---
title: "Global Climate Classification Change Analysis"
subtitle: "Recent Period vs Climate Standard Reference Period"
abstract: "Step 1:  Download ERA5 climate data."
author: Rick Forest
date: today

execute:
  freeze: auto               # only re-run chapters that changed freeze: auto 
  cache: true                # per-chapter caching

format:
  html:
    toc: true




    
---

#### ***Climatological Reference Period Normals***

Average climate computed for a consecutive 30 year period defined by the World Meteorological Organization (WMO)
  <https://community.wmo.int/en/activity-areas/climate-services/climate-products-and-initiatives/wmo-climatological-normals>

```{r climatological-reference-period}
crp_beg <- 1961
crp_end <- 1990
crp_lbl <- paste("CRP ", crp_beg, "-", crp_end, sep = "")
```

#### ***Climate Period Average ***

Average climate computed for a period of at least ten years starting on 1 January of a year ending with the digit 1

```{r climate-period-average}
cpa_beg <- 2011
cpa_end <- 2020
cpa_lbl <- paste("CPA ", cpa_beg, "-", cpa_end, sep = "")
```

#### Packages

```{r setup-packages}
suppressPackageStartupMessages({
  library(ecmwfr)
  library(rnaturalearth)
  library(sf)
  library(dplyr)
  library(gt)
})
```

#### Paths

```{r setup-paths}
dir_download <- "_data/1-download/"
```

#### Download ERA5 climate data

```{r fn-ERA5-data-request}

fn_download_ERA5_data_file <- function(pth, cv, beg, end) {
  yrs <- beg:end
  mths <- 1:12
  if (cv == "t") {
    variable_name <- "2m_temperature"
  } else if (cv == "p") {
    variable_name <- "total_precipitation"
  }
  request <- list(
    product_type = "monthly_averaged_reanalysis",
    dataset_short_name = "reanalysis-era5-single-levels-monthly-means",
    variable = variable_name,
    year = yrs,
    month = mths,
    time = c("00:00"),
    format = "netcdf",
    download_format = "unarchived",
    target = pth
  )
  # Climate Data Store key
  options(keyring_backend = "file") # for ecmwfr to use file instead of mac keychain
  wf_get_key() # keyring password required 
  file <- wf_request(request = request, transfer = TRUE, path = ".")
}
```

```{r fn-download-climate-period-data}

fn_download_cv_data <- function(dir, cp, cv, beg, end) {
  fil <- paste("era5-", cp, "-", beg, "-", end , "-", cv, ".nc", sep = "")
  pth <- paste(dir, fil, sep = "")
  if (!file.exists(pth)) {
    fn_download_ERA5_data_file(pth, cv, beg, end) 
  }
}
```

```{r download-ERA5-climate-data}

dir <- dir_download

cp <- "crp"
cv <- "t"
fn_download_cv_data(dir, cp, cv, crp_beg, crp_end)
cv <- "p"
fn_download_cv_data(dir, cp, cv, crp_beg, crp_end)

cp <- "cpa"
cv <- "t"
fn_download_cv_data(dir, cp, cv, cpa_beg, cpa_end)
cv <- "p"
fn_download_cv_data(dir, cp, cv, cpa_beg, cpa_end)
```

#### Download ERA5 land-sea mask

```{r download-land-sea-mask} 
dir <- dir_download
fil <- "era5-land-sea-mask.nc"
pth <- paste(dir, fil, sep = "")
if (!file.exists(pth))  {
  request <- list(
    product_type = "reanalysis",
    dataset_short_name = "reanalysis-era5-single-levels",
    variable = "land_sea_mask",
    year = "2000", # any date works; the mask is invariant
    month = "01",
    day = "01",
    time = "00:00",
    area = "90/-179.75/-90/180", # N/W/S/E (global)
    grid = "0.25/0.25", # ERA5 native grid spacing
    format = "netcdf",
    download_format = "unarchived",
    target = pth
  )
  # Climate Data Store key
  options(keyring_backend = "file") # for ecmwfr to use file instead of mac keychain
  wf_get_key() # keyring password required 
  file <- wf_request(request = request, transfer = TRUE, path = ".")
}
```

#### Download natural earth coastline

```{r download-ne-coastline}
dir <- dir_download
fil <- "ne_coastline.gpkg"
pth <- paste(dir, fil, sep = "")
if (!file.exists(pth))  {
  ne_coastline_sf <-
    ne_download(
      scale = 10,
      type = "coastline",
      category = "physical",
      returnclass = "sf"
    )
  st_write(ne_coastline_sf, pth, delete_dsn = TRUE, quiet = TRUE)
}
```

#### List of files downloaded

```{r}
dir_path <- dir_download  # change to your directory

files_df <- list.files(dir_path, full.names = TRUE) |>
  file.info() |>
  as.data.frame() |>
  tibble::rownames_to_column("path") |>
  transmute(
    filename = format(basename(path), justify = "left"),
    filesize_MB  = round(size / (1024^2), 2),   # bytes → MB, rounded
    modified_date = as.Date(mtime)
  )

files_df |>
  gt() |>
  cols_align(align = "left", columns = filename) |>
  cols_align(align = "right", columns = filesize_MB) |>
  cols_align(align = "center", columns = modified_date)
```