---
title: "Determinant Calculation"
---

Calculate climate classification determinants using the temperature and precipitation data.

The calculation process is done using **R**.

It will use packages:

  * **tidyverse**
  
    * to calculate the calculate the temparature and precipitation determinants
    * to calculate the aridity threshold
    * to summarize and combine results for each climate period

  * **gt**

    * to produce and display a table of the summarized results


The definitions for the Koppen-Trewartha classification are based on this research:
<https://scispace.com/pdf/climate-classification-revisited-from-koppen-to-trewartha-40ofumhxzu.pdf>

## Initialization

### Periods
```{r}
#| label: periods

# Climatological Reference Period (crp)
crp_beg <- 1961
crp_end <- 1990

# Decadal Climate Period (dcp)
dcp_beg <- 2015
dcp_end <- 2024

```

### Packages
```{r}
#| label: packages
suppressPackageStartupMessages({



  # Primary packages
  library(tidyverse)
  library(gt)

  # Utililty packages
  library(arrow) # to import and export files in parquet format
  library(patchwork) # to arrange multiple plots into one visualization
  library(RColorBrewer)
  library(scales)
})
```

### Directories
```{r}
#| label: directories
climate_aggregation_dir <- "data/02_climate_aggregation/"
determinant_calculation_dir <- "data/03_determinant_calculation/"
```

## Functions

```{r}
#| label: fn_weighted_average
fn_weighted_average <- function(df) {

   metric_order <- c("tavg", "tmax", "tmin", "tg10", "ptot",
                     "pwnt", "psmr", "pmin", "pl60", "athr")

   df_wgtavg  <- df %>%
    select(area,tavg,tmax,tmin,tg10,
           ptot,pmin,pl60,pwnt,psmr,athr) |>
    pivot_longer(
      cols = c(tavg,tmax,tmin,tg10,ptot,pmin,pl60,pwnt,psmr,athr),
      names_to = "metric",
      values_to = "value"
    ) |>
    group_by(metric) %>%
    summarise(
      wgtavg = weighted.mean(value, w = area, na.rm = TRUE),
      .groups = "drop"
    )

    df_wgtavg <- df_wgtavg |>
      mutate(metric = factor(metric, levels = metric_order)) %>%
      arrange(metric)

    return(df_wgtavg)
}
```

## Import
```{r}
#| label: import

fil = paste0("t_agg_", crp_beg, "-", crp_end, ".parquet")
dir <- climate_aggregation_dir 
pth <- paste0(dir, fil)
crp_t_df <- read_parquet(pth) 

fil = paste0("p_agg_", crp_beg, "-", crp_end, ".parquet")
dir <- climate_aggregation_dir 
pth <- paste0(dir, fil)
crp_p_df <- read_parquet(pth) 

fil = paste0("t_agg_", dcp_beg, "-", dcp_end, ".parquet")
dir <- climate_aggregation_dir 
pth <- paste0(dir, fil)
dcp_t_df <- read_parquet(pth) 

fil = paste0("p_agg_", dcp_beg, "-", dcp_end, ".parquet")
dir <- climate_aggregation_dir 
pth <- paste0(dir, fil)
dcp_p_df <- read_parquet(pth) 
```

## Temperature Determinants

### Function

```{r}
#| label: fn_t_determinants 
fn_t_determinants <- function(df) {

  mcols <- paste0("m", 1:12)
  df <- df |>
    filter(lsm > 0) |>
    mutate(
        tavg = (m1 + m2 + m3 + m4 + m5 + m6 + m7 + m8 + m9 + m10 + m11 + m12) / 12,
        tmax = pmax(m1, m2, m3, m4, m5, m6, m7, m8, m9, m10, m11, m12),
        tmin = pmin(m1, m2, m3, m4, m5, m6, m7, m8, m9, m10, m11, m12),
        tg10 = rowSums(across(all_of(mcols), ~ .x >= 10)),
        tg10 = as.integer(tg10)
    )  
  df <- df |>
    select(-m1, -m2, -m3, -m4, -m5, -m6, -m7, -m8, -m9, -m10, -m11, -m12)
  
  return(df)
}
```

### Calculate

```{r}
#| label: calc_t
crp_t_det_df <- fn_t_determinants(crp_t_df)
dcp_t_det_df <- fn_t_determinants(dcp_t_df)
```



### Determinant: Average Monthly Temperature

Average Monthly Temperature is used to in the aridity threshold calculation.  

```{r}
#| label: t1
#| fig-width: 10
#| fig-aspect: .7   
#| out-width: 100%
 
df1 <- crp_t_det_df |> select(lon, lat, area, tavg) 
df2 <- dcp_t_det_df |> select(lon, lat, area, tavg) 
df <- df1 |>
  left_join(df2, by = c("lon", "lat", "area")) |>
  select(tavg.x, tavg.y, area) |>
  rename(CRP = tavg.x, DCP = tavg.y) |>
  pivot_longer(cols = c(CRP, DCP), names_to  = "cp", values_to = "tavg")

ggplot(df, aes(x = tavg, weight = area, fill = ifelse(tavg <=0, "cold", "warm"))) +
  geom_histogram(binwidth = 2, boundary = 10, color = "white") + 
  scale_fill_manual(
    values = c(
      cold = "#7fb5db",               
      warm = "#c87070"), 
    name = "Category") +
  scale_x_continuous(breaks = seq(-40, ceiling(max(df$tavg)), by = 10)) +
  labs(
    title = "Average Monthly Temperature",
    subtitle = "Used to Calculate Aridity Threshold",
    x = "°C",
    y = "Weighted Count (by Area)") +
  facet_wrap(~ cp, ncol = 1) +   # one panel for CRP, one for DCP
  theme_minimal(base_size = 14) +
  guides(fill = "none")

```

### Determinant: Maximum Month Temperature

Maximum Average Monthly Temperature is used to identify cold regions, using these thresholds:  

  * Ice Cap: < 0 °C
  * Tundra : 0°C to 10°C

```{r}
#| label: t2
#| fig-width: 10
#| fig-aspect: .7   
#| out-width: 100%
 
df1 <- crp_t_det_df |> select(lon, lat, area, tmax) 
df2 <- dcp_t_det_df |> select(lon, lat, area, tmax) 
df <- df1 |>
  left_join(df2, by = c("lon", "lat", "area")) |>
  select(tmax.x, tmax.y, area) |>
  rename(CRP = tmax.x, DCP = tmax.y) |>
  pivot_longer(cols = c(CRP, DCP), names_to  = "cp", values_to = "tmax") |>
  mutate(
    category = factor(
      case_when(
        tmax < 0   ~ "Ice Cap",
        tmax < 10  ~ "Tundra",
        tmax >= 10 ~ "Other"
      ),
      levels = c("Ice Cap", "Tundra", "Other")
    )
  )

ggplot(df, aes(x = tmax, weight = area, fill = category)) +
  geom_histogram(binwidth = 2, boundary = 10, color = "white") +  # 10 °C is a bin edge
  scale_fill_manual(
    values = c("Ice Cap" = "#27679a",   # cool blue-teal
               "Tundra" = "#7fb5db",  # lighter blue-teal
               "Other" = "#d9d9d9"),  # neutral grey
    name = "Category") +
  scale_x_continuous(breaks = seq(-40, ceiling(max(df$tmax)), by = 10)) +
  labs(
    title = "Maximum Average Monthly Temperature < 10°C",
    subtitle = "Determines Cold Climates",
    x = "°C",
    y = "Weighted Count (by Area)") +
  facet_wrap(~ cp, ncol = 1) +   # one panel for CRP, one for DCP
  theme_minimal(base_size = 14) +
  guides(fill = "none")
```

### Determinant: Minimum Month Temperature

Lowest Monthly Average Temperature is used to identify hot regions, using this threshold:  

  * Tropical: > 18°C

```{r}
#| label: t3
#| fig-width: 10
#| fig-aspect: .7   
#| out-width: 100%

df1 <- crp_t_det_df |> select(lon, lat, area, tmin) 
df2 <- dcp_t_det_df |> select(lon, lat, area, tmin) 
df <- df1 |>
  left_join(df2, by = c("lon", "lat", "area")) |>
  select(tmin.x, tmin.y, area) |>
  rename(CRP = tmin.x, DCP = tmin.y) |>
  pivot_longer(cols = c(CRP, DCP), names_to  = "cp", values_to = "tmin")

ggplot(df, aes(x = tmin, weight = area, fill = ifelse(tmin > 18, "hot", "other"))) +
  geom_histogram(binwidth = 2, boundary = 18, color = "white") +  
  scale_fill_manual(
    values = c("hot" = "#8b0000",  
               "other" = "#d9d9d9"),  # neutral grey
  ) +
  scale_x_continuous(
    breaks = seq(0, max(df$tmin, na.rm = TRUE), by = 6)
  ) +
  labs(
    title = "Minimum Temperature > 18°C",
    subtitle = "Determines Hot Climates",
    x = "°C",
    y = "Weighted Count (by Area)"
  ) +
  facet_wrap(~ cp, ncol = 1) +   # one panel for CRP, one for DCP
  theme_minimal(base_size = 14) +
  guides(fill = "none")
```

### Determinant: Months > 10°C

The count of months with average temperature > 10°C is used to identify subtropical, temperate, and boreal climates, using these thresholds: 

  * Subtropical: > 8 or more months
  * Temperate: 4 to 7 months
  * Boreal: 1 to 3 months

```{r}
#| label: t4
#| fig-width: 10
#| fig-aspect: .7   
#| out-width: 100%

df1 <- crp_t_det_df |>
  filter(lsm >0 ) |>
  select(lon, lat, area, tg10) 
df2 <- dcp_t_det_df |>
  filter(lsm >0 ) |>
  select(lon, lat, area, tg10) 

df <- df1 |>
  left_join(df2, by = c("lon", "lat", "area")) |>
  select(tg10.x, tg10.y, area) |>
  rename(CRP = tg10.x,
         DCP = tg10.y)

bucket <- function(x) case_when(
  x >= 8 ~ "8-12",
  x >= 4 ~ "4-7",
  x >= 1 ~ "1-3",
  x == 0 ~ "0"
)

df_long <- df %>%
  mutate(
    cat_crp = bucket(CRP),
    cat_dcp = bucket(DCP)
  ) %>%
  pivot_longer(
    cols = c(cat_crp, cat_dcp),
    names_to = "period",
    values_to = "category"
  ) %>%
  group_by(period, category) %>%
  summarise(total_area = sum(area, na.rm = TRUE), .groups = "drop") %>%
  mutate(period = recode(period, cat_crp = "CRP", cat_dcp = "DCP")) %>%
  # compute proportion of grand total
  mutate(prop_area = total_area / sum(total_area))

df_long$category <- factor(df_long$category,
                           levels = c("0", "1-3", "4-7", "8-12"))

ggplot(df_long,
       aes(x = category, y = prop_area, fill = period)) +
  geom_col(position = "dodge", width = 0.7) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_fill_brewer(palette = "Dark2") + # or "Dark2", "Paired", etc.
  labs(
    title = "Classification Determinant: Temperature",
    subtitle = "Months with Temperature > 10°C",
    x = "Months per Year",
    y = "Land Area %",
    fill = "Climate\nPeriod",
  ) +
  theme_minimal(base_size = 14)

```

## Precipitation Determinants

### Function

```{r}
#| label: fn_p_determinants 
fn_p_determinants <- function(df) {

  mcols <- paste0("m", 1:12)
  df <- df |>
    filter(lsm > 0) |>
    mutate(
      ptot = (m1 + m2 + m3 + m4 + m5 + m6 + m7 + m8 + m9 + m10 + m11 + m12),
      pmin = pmin(m1, m2, m3, m4, m5, m6, m7, m8, m9, m10, m11, m12),
      pl60 = rowSums(across(all_of(mcols), ~ .x < 60)),
      pl60 = as.integer(pl60),
      pwnt = case_when(
        lat >= 0 ~ (m1 + m2 + m3 + m10 + m11 + m12),
        lat < 0 ~ (m4 + m5 + m6 + m7 + m8 + m9)
      ),
      psmr = case_when(
        lat >= 0 ~ (m4 + m5 + m6 + m7 + m8 + m9),
        lat < 0 ~  (m1 + m2 + m3 + m10 + m11 + m12)
      )
    )     
  df <- df |>
    select(-m1, -m2, -m3, -m4, -m5, -m6, -m7, -m8, -m9, -m10, -m11, -m12)

  return(df)
}
```

### Calculate

```{r}
#| label: calc_p
crp_p_det_df <- fn_p_determinants(crp_p_df)
dcp_p_det_df <- fn_p_determinants(dcp_p_df)
```

### Determinant: Seasonal Precipitation

Total Precipitation will be used to identify dry regions.  These are the climates identified by this determinant:

  * Desert (Arid)     : < Aridity Threshold / 2
  * Steppe (Semiarid) : < Aridity Threshold

```{r}
#| label: p1
#| fig-width: 10
#| fig-aspect: .7   
#| out-width: 100%

df1 <- crp_p_det_df |>
  filter(lsm > 0 ) |>
  select(lon, lat, ptot, pwnt, psmr, area) 
df2 <- dcp_p_det_df  |>
  filter(lsm > 0 ) |>
  select(lon, lat, ptot, pwnt, psmr, area) 
dfj <- df1 |>
  left_join(df2, by = c("lon", "lat", "area")) |>
  select(ptot.x, ptot.y, pwnt.x, pwnt.y, psmr.x, psmr.y, area)

df_long <- dfj %>%
  pivot_longer(
    cols = c(ptot.x, ptot.y, pwnt.x, pwnt.y, psmr.x, psmr.y),
    names_to = c("metric", "period"),
    names_pattern = "^p(.*)\\.(x|y)$",
    values_to = "precip"
  ) %>%
  mutate(
    metric = recode(metric,
                    smr = "Summer",
                    wnt = "Winter",
                    tot = "Total"),
    period = recode(period,
                    x = "CRP",
                    y = "DCP")
  )

# area-weighted mean per metric × period
summ <- df_long %>%
  group_by(metric, period) %>%
  summarise(
    mean_precip = weighted.mean(precip, w = area, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(summ, aes(x = period, y = mean_precip, fill = period)) +
  geom_bar(stat = "identity", position = "dodge") +
  facet_wrap(~ metric, ncol = 3) +  
  labs(
    title = "Seasonal Precipitation",
    x = NULL,
    y = "Precipitation (mm)",
    fill = "period"
  ) +
  scale_fill_brewer(palette = "Paired") +  # or "Dark2", "Paired", etc.
  theme_minimal(base_size = 14) +
  theme(
    #legend.position = "none",
    #strip.text = element_text(face = "bold"),
    axis.text.x = element_blank(),                  # hide tick labels if desired
    axis.ticks.x = element_blank()
  )

```

## Aridity Threshold

### Function

```{r}
#| label: fn_aridity_threshold

fn_aridity_threshold <- function(df) {

# Aridity threshold formula: R = 2.3T – 0.64Pw + 41 

    # R: mean annual precipitation threshold in cm

    # T: mean annual air temperature in °C
    # Pw: percentage of annual precipitation in winter (6 low-sun months)
    # winter: October to March in Northern Hemisphere
    #         April to September in Southern Hemisphere

  df <- df |>
    mutate(
      term1 = (2.3 * tavg),
      term2 = 
        if_else(ptot != 0, 
          (0.64 * ((pwnt / ptot) * 100)), # multiply by 100 to convert to %
          0),
      term3 = 41,
      athr = (term1 - term2 + term3) * 10 # convert result from cm to mm

    )  |>
    select(-term1, -term2, -term3)

  return(df)
}
```

### Calculate

```{r}
#| label: calc_at
crp_j_df <- crp_t_det_df |>
  left_join(crp_p_det_df, by = c("lon", "lat", "area", "lsm"))
crp_at_df <- fn_aridity_threshold(crp_j_df)

dcp_j_df <- dcp_t_det_df |>
  left_join(dcp_p_det_df,, by = c("lon", "lat", "area", "lsm"))
dcp_at_df <- fn_aridity_threshold(dcp_j_df)
```

### Determinant: Aridity Threshold

```{r}
#| label: at1
#| fig-width: 10
#| fig-aspect: .7   
#| out-width: 100%

df1 <- crp_at_df |>
  filter(lsm > 0 ) |>
  mutate(Area = area * lsm) |>
  mutate(
    Climate = factor(
      case_when(
        ptot < (athr/2) ~ "Arid (TP < AT/2)",
        ptot <= athr    ~ "Semi-Arid (TP <= AT",
        TRUE            ~ "Not Dry (Pt >= AT)"),
      levels = c(
        "Arid (TP < AT/2)", 
        "Semi-Arid (TP <= AT", 
        "Not Dry (Pt >= AT)"), 
    )
  ) |>
  select(lon, lat, ptot, Area, Climate) |>
  group_by(Climate) |>
  summarize(sum_area=sum(Area), .groups = "drop")

df2 <- dcp_at_df  |>
  filter(lsm > 0 ) |>
  mutate(Area = area * lsm) |>
  mutate(
    Climate = factor(
      case_when(
        ptot < (athr/2) ~ "Arid (TP < AT/2)",
        ptot <= athr    ~ "Semi-Arid (TP <= AT",
        TRUE            ~ "Not Dry (Pt >= AT)"),
      levels = c(
        "Arid (TP < AT/2)", 
        "Semi-Arid (TP <= AT", 
        "Not Dry (Pt >= AT)"), 
    )
  ) |>
  select(lon, lat, ptot, Area, Climate) |>
  group_by(Climate) |>
  summarize(sum_area=sum(Area), .groups = "drop")

dfj <- df1 |>
  left_join(df2, by = c("Climate")) |>
  rename(CRP = sum_area.x,
         DCP = sum_area.y) |>
  pivot_longer(
    cols = c(CRP, DCP),
    names_to  = "Period",
    values_to = "Area"
  )

dfj_pct <- dfj |>
  group_by(Period) |>
  mutate(pct_area = Area / sum(Area) * 100)


ggplot(dfj_pct, 
  aes(x = Climate, y = pct_area, fill = Period)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_brewer(palette = "Paired") +  # or "Dark2", "Paired", etc.
  labs(
    title = "Total Precipitation (TP) vs Aridity Threshold (AT)",
    subtitle = "Identifies Dry Climates",
    x = "",
    y = "% of Total Area",
    fill = "Period"
  ) +
  theme_minimal(base_size = 14)

```

## Save Results

```{r}
#| label: save_results

pth = paste0(determinant_calculation_dir, "crp_determinants.parquet")
write_parquet(crp_at_df, pth)

pth = paste0(determinant_calculation_dir, "dcp_determinants.parquet")
write_parquet(dcp_at_df, pth)

```



## Build Summary Table

```{r}
#| label: build_table
df1 <- fn_weighted_average(crp_at_df)
df2 <- fn_weighted_average(dcp_at_df)

df_wgtavg <- df1 |> 
  full_join(df2, by = "metric") |> 
  rename(
    crp_metric = wgtavg.x,
    dcp_metric = wgtavg.y) |> 
  mutate(metric_chg = dcp_metric - crp_metric) |> 
  mutate(pct_chg = metric_chg / crp_metric) |> 
  mutate(
    metric = recode(metric,
      "tavg" = "Temperature: Avg",
      "tmax" = "Temperature: Max",
      "tmin" = "Temperature: Min",
      "tg10" = "Temperature: Months > 10°C",
      "ptot" = "Precipitation: Total",
      "pmin" = "Precipitation: Min",
      "pl60" = "Precipitation: Months < 60mm",
      "pwnt" = "Precipitation: Winter",
      "psmr" = "Precipitation: Summer",
      "athr" = "Aridity Threshold"
    )
  )


```

## Display Summary Table

```{r}
#| label: display_table
rows_to_style <- which(
  df_wgtavg[["pct_chg"]] > 0.05)

gt_tbl <- 
  df_wgtavg |>
    gt(rowname_col = "metric") |>
    tab_header(
      title = "Climate Classification Determinants Summary",
      subtitle = "Area-Weighted Averages"
    ) |>
    cols_label(
      crp_metric = "CRP",
      dcp_metric = "DCP",
      metric_chg = "Change",
      pct_chg = "% Change",
    ) |>
    fmt_number(
      columns = c(crp_metric, dcp_metric, metric_chg, pct_chg),
      decimals = 2
    ) |>
    tab_style(
      style = cell_fill(color = "#e6f0fa"),          # highlight color
      locations = cells_body(columns = "pct_chg", rows = rows_to_style)
    ) |>
    tab_style(
      style = cell_text(align = "left"),   # left-justify
      locations = cells_stub()
    )


gt_tbl

```


