---
title: "Determinant Calculation"
format:
  html:
    toc: true
    number-sections: false
---

Calculate climate classification determinants (temperature, precipitation, aridity) for two periods and summarize/visualize area-weighted results. These feed a later step that applies Köppen–Trewartha classification.

**Determinants computed**

- Temperature: `tavg` (annual average), `tmax` (max monthly), `tmin` (min monthly), `tg10` (months > 10 °C)
- Precipitation: `ptot` (annual total), `pwnt` (winter total), `psmr` (summer total), `pmin` (min monthly), `pl60` (months < 60 mm)
- Aridity Threshold: `athr` (calculated using temperature and precipitation determinants)

The classification definitions are based on  
<https://scispace.com/pdf/climate-classification-revisited-from-koppen-to-trewartha-40ofumhxzu.pdf>

## Initialization

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, cache = TRUE)
```

### Packages
```{r}
#| label: packages
suppressPackageStartupMessages({
  library(tidyverse)
  library(gt)
  library(arrow)
  library(patchwork)
  library(scales)
})
```

### Periods
```{r}
#| label: periods
periods <- list(
  crp = list(beg = 1961, end = 1990),
  dcp = list(beg = 2015, end = 2024)
)
CRP_YEARS <- c(periods$crp$beg, periods$crp$end)
DCP_YEARS <- c(periods$dcp$beg, periods$dcp$end)

CRP_LBL <- sprintf("CRP (%d–%d)", CRP_YEARS[1], CRP_YEARS[2])
DCP_LBL <- sprintf("DCP (%d–%d)", DCP_YEARS[1], DCP_YEARS[2])

```

### Directories
```{r}
#| label: directories
idir <- "data/02_climate_aggregation/"
odir <- "data/03_determinant_calculation/"
dir.create(odir, recursive = TRUE, showWarnings = FALSE)
```

### Metrics
```{r}
#| label: metrics
temp_metrics   <- c("tavg", "tmax", "tmin", "tg10")
precip_metrics <- c("ptot", "pwnt", "psmr", "pmin", "pl60")
```

### Palette
```{r}
#| label: palette
ktc_palette <- list(
  tropical    = "#c23734",  
  dry         = "#f09758",   
  subtropical = "#f28e2b",   
  polar       = "#1552a2",   
  boreal      = "#2b74cc",  
  temperate   = "#3CB371",  
  other       = "#e0e0e0",  
  rainforest  = "#02745E",   # deep green
  savanna     = "#DDAA33"    # gold (color-blind friendly)
 )

# Define default colors for all plots
role_palette <- list(
  period  = c(
    CRP = "#7B66D2",
    DCP = "#009E73"),    
  season  = c(
    Winter = "#1552a2",
    Summer = "#c23734"),
  dryness = c(
    Dry = "#f09758",
    Humid = "#1552a2")
)
```

### Seasons
```{r}
#| label: seasons
NORTH_WINTER <- paste0("m", c(1, 2, 3, 10, 11, 12))
NORTH_SUMMER <- paste0("m", 4:9)
```

### Theme
```{r}
#| label: theme
theme_custom <- theme_minimal(base_size = 14)
theme_set(theme_custom)
```

### Helpers
```{r}
#| label: helpers

# Period labeling
pretty_period_lbl <- function(x) recode(x, CRP = CRP_LBL, DCP = DCP_LBL)

# Area-weighted average for a set of metrics for CRP/DCP lists
fn_weighted_average <- function(df, metrics) {
  df |>
    select(area, any_of(metrics)) |>
    pivot_longer(-area, names_to = "metric", values_to = "value") |>
    group_by(metric) |>
    summarise(wgtavg = weighted.mean(value, w = area, na.rm = TRUE), .groups = "drop")
}

wgt_by_metric <- function(det_list, metrics) {
  bind_rows(
    fn_weighted_average(det_list$crp, metrics) |> mutate(period = "CRP"),
    fn_weighted_average(det_list$dcp, metrics) |> mutate(period = "DCP")
  ) |>
    pivot_wider(names_from = period, values_from = wgtavg) |>
    rename(crp_metric = CRP, dcp_metric = DCP) |>
    mutate(metric_chg = dcp_metric - crp_metric)
}

# Long-to-tidy two-period prep
prepare_period_df <- function(df_crp, df_dcp, vars) {
  df_crp |>
    select(lon, lat, area, all_of(vars)) |>
    left_join(
      df_dcp |> select(lon, lat, area, all_of(vars)),
      by = c("lon", "lat") # safer join; area should match but don't join on it
    ) |>
    pivot_longer(
      cols = ends_with(c(".x", ".y")),
      names_to = c(".value", "period_side"),
      names_sep = "\\."
    ) |>
    transmute(
      lon, lat, area,
      across(all_of(vars)),
      period = if_else(period_side == "x", "CRP", "DCP"),
      period_lbl = factor(pretty_period_lbl(period),levels = c(CRP_LBL, DCP_LBL))
    )
}

# Reusable gt table helper for CRP/DCP + Change
gt_det_table <- function(df, title, subtitle, decimals = 2) {
  df %>%
    gt(rowname_col = "metric") %>%
    tab_header(title = title, subtitle = subtitle) %>%
    cols_label(crp_metric = "CRP", dcp_metric = "DCP", metric_chg = "Change") %>%
    fmt_number(columns = c(crp_metric, dcp_metric, metric_chg), decimals = decimals) %>%
    tab_style(style = cell_text(align = "left"),  locations = cells_stub()) %>%
    tab_style(style = cell_text(v_align = "middle"), locations = cells_body(rows = everything())) %>%
    tab_style(style = cell_text(align = "right"),
              locations = cells_body(columns = c(crp_metric, dcp_metric, metric_chg))) %>%
    tab_style(style = cell_text(align = "center"),
              locations = cells_column_labels(columns = c(crp_metric, dcp_metric, metric_chg)))
}
```

## Import Data
```{r}
#| label: import
import_data <- function(type, period_name) {
  beg <- periods[[period_name]]$beg
  end <- periods[[period_name]]$end
  fil <- paste0(type, "_agg_", beg, "-", end, ".parquet")
  read_parquet(file.path(idir, fil))
}
t_import <- list(crp = import_data("t", "crp"),
                 dcp = import_data("t", "dcp"))
p_import <- list(crp = import_data("p", "crp"),
                 dcp = import_data("p", "dcp"))
```

## Calculate Determinants

Temperature determinants (CRP vs DCP). Area‑weighted averages for tavg, tmax, tmin, and months > 10 °C (TG10). Shifts indicate warming/cooling and potential movement across Köppen–Trewartha thermal thresholds (10 °C, 18 °C).

```{r}
#| label: t_determinants

calculate_t_determinants <- function(df) {
  mcols <- paste0("m", 1:12)
  df %>%
    filter(lsm > 0) %>%
    mutate(
      # average across 12 monthly columns
      tavg = rowMeans(across(all_of(mcols)), na.rm = TRUE),
      # row-wise max/min via do.call on the across() columns
      tmax = do.call(pmax, c(across(all_of(mcols)), na.rm = TRUE)),
      tmin = do.call(pmin, c(across(all_of(mcols)), na.rm = TRUE)),
      # count months >= 10 °C
      tg10 = rowSums(across(all_of(mcols), ~ .x >= 10), na.rm = TRUE)
    ) %>%
    select(-all_of(mcols))
}

t_det <- map(t_import, calculate_t_determinants)

df_wgtavg_temp <- wgt_by_metric(t_det, temp_metrics) |>
  mutate(metric = recode(metric,
    tavg = "Temperature: Avg",
    tmax = "Temperature: Max",
    tmin = "Temperature: Min",
    tg10 = "Temperature: Months > 10°C"
  )) |>
  arrange(factor(metric, c(
    "Temperature: Avg","Temperature: Max","Temperature: Min","Temperature: Months > 10°C"
  )))

gt_tbl_temp <- gt_det_table(
  df_wgtavg_temp,
  title    = "Temperature Determinants Summary",
  subtitle = "Area-Weighted Averages (°C)",
  decimals = 2
)

gt_tbl_temp
```

Precipitation determinants (CRP vs DCP). Area‑weighted averages for annual total (ptot), winter/summer share (pwnt/psmr), driest month (pmin), and months < 60 mm (PL60). Changes influence dry/wet regimes and tropical subtype splits.

```{r}
#| label: p_determinants

calculate_p_determinants <- function(df) {
  mcols <- paste0("m", 1:12)
  df %>%
    filter(lsm > 0) %>%
    mutate(
      ptot = rowSums(across(all_of(mcols)), na.rm = TRUE),
      pmin = do.call(pmin, c(across(all_of(mcols)), na.rm = TRUE)),
      pl60 = rowSums(across(all_of(mcols), ~ .x < 60), na.rm = TRUE),
      pwnt = if_else(
        lat >= 0,
        rowSums(across(all_of(NORTH_WINTER)), na.rm = TRUE),
        rowSums(across(all_of(NORTH_SUMMER)), na.rm = TRUE)
      ),
      psmr = if_else(
        lat >= 0,
        rowSums(across(all_of(NORTH_SUMMER)), na.rm = TRUE),
        rowSums(across(all_of(NORTH_WINTER)), na.rm = TRUE)
      )
    ) %>%
    select(-all_of(mcols))
}

p_det <- map(p_import, calculate_p_determinants)

df_wgtavg_precip <- wgt_by_metric(p_det, precip_metrics) |>
  mutate(metric = recode(metric,
    ptot = "Precipitation: Total",
    pmin = "Precipitation: Min",
    pl60 = "Precipitation: Months < 60 mm",
    pwnt = "Precipitation: Winter",
    psmr = "Precipitation: Summer"
  )) |>
  arrange(factor(metric, c(
    "Precipitation: Total",
    "Precipitation: Winter",
    "Precipitation: Summer",
    "Precipitation: Min",
    "Precipitation: Months < 60 mm"
  )))

gt_tbl_precip <- gt_det_table(
  df_wgtavg_precip,
  title    = "Precipitation Determinants Summary",
  subtitle = "Area-Weighted Averages (mm)",
  decimals = 2
)

gt_tbl_precip
```

Aridity components and threshold. Decomposes the aridity formula (2.3 × tavg) + (−0.64 × winter %) + 41 into terms and the resulting threshold (athr). A higher athr increases the precipation needed to not be classified as dry.

```{r}
#| label: a_determinant
calculate_aridity_threshold <- function(df) {
  df %>%
    mutate(
      pw_pct = if_else(ptot != 0, (pwnt / ptot) * 100, 0),
      athr   = (2.3 * tavg - 0.64 * pw_pct + 41) * 10  # cm to mm -> mm
    ) %>%
    select(-pw_pct)
}

# Build aridity inputs with a safe join and unified columns
a_det <- map2(
  t_det, p_det,
  ~ left_join(.x, .y, by = c("lon","lat"), suffix = c(".t",".p")) %>%
      mutate(
        # unify duplicates from temperature (.t) and precipitation (.p)
        area = coalesce(area.t, area.p),
        lsm  = coalesce(lsm.t,  lsm.p)
      ) %>%
      select(-ends_with(".t"), -ends_with(".p")) %>%
      calculate_aridity_threshold()
)

# Robust weighted-term calculator (ensures x and w have same length)
calc_weighted_term <- function(df, term_name, expr) {
  df %>%
    transmute(area, value = {{ expr }}) %>%
    filter(is.finite(area), area > 0, is.finite(value)) %>%
    summarise(wgtavg = weighted.mean(value, w = area, na.rm = TRUE)) %>%
    mutate(metric = term_name) %>%
    select(metric, wgtavg)
}

# Compute weighted averages for the three terms in each period
df_wgtavg_aridity <- map2_df(
  rep(list(a_det$crp, a_det$dcp), each = 3),
  rep(c("tavg_term1", "pwnt_ptot", "athr"), 2),
  ~ calc_weighted_term(
      .x,
      .y,
      switch(.y,
        tavg_term1 = tavg * 2.3,
        pwnt_ptot  = if_else(ptot != 0, ((pwnt / ptot) * 100 * -0.64) + 41, 41),
        athr       = athr
      )
    )
) %>%
  mutate(period = rep(c("crp", "dcp"), each = 3)) %>%
  pivot_wider(names_from = period, values_from = wgtavg, names_prefix = "wgtavg_") %>%
  rename(crp_metric = wgtavg_crp, dcp_metric = wgtavg_dcp) %>%
  mutate(
    metric_chg = dcp_metric - crp_metric,
    metric = recode(metric,
      tavg_term1 = "Term1: Annual Average Temperature * 2.3",
      pwnt_ptot  = "Term2: (% of Precipitation in Winter) * -0.64 + 41",
      athr       = "Aridity Threshold: (Term1 + Term2) * 10"
    ),
    metric = factor(metric, levels = c(
      "Term1: Annual Average Temperature * 2.3",
      "Term2: (% of Precipitation in Winter) * -0.64 + 41",
      "Aridity Threshold: (Term1 + Term2) * 10"
    ))
  ) %>%
  arrange(metric)


gt_tbl_aridity <- gt_det_table(
  df_wgtavg_aridity,
  title    = "Aridity Threshold Summary",
  subtitle = "Area-Weighted Averages (mm)",
  decimals = 2
)

gt_tbl_aridity
```

## Visualizations

### Temperature 



```{r}
#| label: v_t_hist_prep
temp_df <- prepare_period_df(a_det$crp, a_det$dcp, c("tavg","tmax","tmin","tg10"))
```

Average monthly temperature distribution. Shows the CRP vs DCP distribution of tavg. A rightward shift implies general warming, increasing Term 1 of the aridity formula.

```{r}
#| label: v_t_avg
p_tavg <- ggplot(temp_df, aes(x = tavg, weight = area, fill = period)) +
  geom_histogram(binwidth = 2, boundary = 10, color = "white") +
  scale_fill_manual(values = role_palette$period) +
  scale_x_continuous(breaks = seq(-40, ceiling(max(temp_df$tavg)), by = 10)) +
  labs(title = "Average Monthly Temperature",
       subtitle = "Used in Aridity Threshold Calculation",
       x = "°C", y = "Weighted Count (by Area)") +
  facet_wrap(~ period_lbl, ncol = 1) +
  theme(legend.position = "none")
```

Average maximum monthly temperature distribution. Shows the CRP vs DCP distribution of tmax. A rightward shift indicates a loss of polar climate zones.

```{r}
#| label: v_t_max
p_tmax <- temp_df |>
  mutate(category = factor(case_when(
    tmax < 10 ~ "Polar",
    TRUE      ~ "Other"
  ), levels = c("Polar","Other"))) |>
  ggplot(aes(x = tmax, weight = area, fill = category)) +
  geom_histogram(binwidth = 2, boundary = 10, color = "white") +
  scale_fill_manual(values = c(Polar = ktc_palette$polar, Other = ktc_palette$other)) +
  scale_x_continuous(breaks = seq(-40, ceiling(max(temp_df$tmax)), by = 10)) +
  labs(title = "Maximum Average Monthly Temperature",
       subtitle = "Identifies polar climates (< 10 °C)",
       x = "°C",
       y = "Weighted Count (by Area)",
       fill = NULL) +
  facet_wrap(~ period_lbl, ncol = 1) +
  theme(legend.position = "top")
```

Average minimum monthly temperature distribution. Shows the CRP vs DCP distribution of tmin. A rightward shift indicates an increase in tropical climate zones.

```{r}
#| label: v_t_min
p_tmin <- temp_df |>
  mutate(category = factor(case_when(
    tmin > 18 ~ "Tropical",
    TRUE      ~ "Other"
  ), levels = c("Tropical","Other"))) |>
  ggplot(aes(x = tmin, weight = area, fill = category)) +
    geom_histogram(binwidth = 2, boundary = 18, color = "white") +
    scale_fill_manual(values = c(Tropical = ktc_palette$tropical, Other = ktc_palette$other)) +
    scale_x_continuous(breaks = seq(0, max(temp_df$tmin, na.rm = TRUE), by = 6)) +
    labs(title = "Minimum Average Monthly Temperature",
         subtitle = "Identifies tropical climates (> 18 °C)",
         x = "°C",
         y = "Weighted Count (by Area)",
         fill = NULL) +
    facet_wrap(~ period_lbl, ncol = 1) +
    theme(legend.position = "top")
```

```{r}
#| label: t_patchwork
#| fig-height: 15
#| fig-aspect: .7
#| out-width: 100%
(p_tavg / p_tmax / p_tmin)
```

```{r}
#| label: v_t_profile_prep
df <- prepare_period_df(a_det$crp, a_det$dcp, c("tmin","tmax","ptot","athr","tg10"))
```

Count of months per year that have average temperature greater than 10 °C. 

```{r}
#| label: t_profile
#| fig-height: 5
#| fig-aspect: .7
#| out-width: 100%

plot_data <- df |>
  filter(tmax > 10, tmin < 18, ptot > athr) |>
  select(period_lbl, tg10, area) |>
  group_by(period_lbl, tg10)  |>
  summarise(total_area = sum(area, na.rm = TRUE), .groups = "drop")  |>
  group_by(period_lbl) |>
  mutate(prop_area = total_area / sum(total_area)) |>
  mutate(category = factor(case_when(
    tg10 >= 8 ~ "Subtropical",
    tg10 >= 4 ~ "Temperate",
    tg10 >= 1 ~ "Boreal",
    TRUE      ~ "Other"
  ), levels = c("Boreal","Temperate","Subtropical","Other")))

p_tg10 <- plot_data |>
  ggplot(aes(x = tg10, y = prop_area, fill = category)) +
  geom_col(position = "dodge", width = 0.7) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  scale_x_continuous(breaks = seq(from = 1, to = 12, by = 1)) +
  scale_fill_manual(
    values = c(
      Boreal      = ktc_palette$boreal,      # cold blue
      Temperate   = ktc_palette$temperate,   # cool cyan/teal
      Subtropical = ktc_palette$subtropical, # warm amber
      Other       = ktc_palette$other)) +
  labs(title = "Months in Year with Average Temperature > 10 °C",
       subtitle = "Identifies Boreal, Temperate and Subtropical Climates",
       x = "Month Count",
       y = "Area %",
       fill = NULL)+
  facet_wrap(~ period_lbl, ncol = 1) +
  theme(legend.position = "top")

p_tg10
```

### Precipitation

Annual precipitation distribution (log10). Comparing CRP vs DCP reveals whether total precipitation shifted. Lower values push more cells toward dryness in the aridity test.

```{r}
#| label: v_p_total
#| fig-height: 5
#| out-width: 100%

df_ptot <- prepare_period_df(a_det$crp, a_det$dcp, c("ptot")) %>%
filter(is.finite(ptot), ptot > 0, area > 0) %>%
mutate(period = factor(period, levels = c("CRP","DCP")),
period_lbl = factor(period_lbl, levels = c(CRP_LBL, DCP_LBL)))

ggplot(df_ptot, aes(x = ptot, weight = area, fill = period)) +
    geom_density(adjust = 1, alpha = 1.0, color = NA) +
    scale_x_log10(labels = label_number(scale_cut = cut_short_scale())) +
    scale_fill_manual(values = role_palette$period) +
    labs(
      title = "Annual Total Precipitation",
      subtitle = "Identifies Dry Climates when < Aridity Threshold",
      x = "mm (log10)",
      y = "Area-weighted density") +
    facet_wrap(~ period_lbl, ncol = 1) +
    theme(legend.position = "none")

```

```{r}
#| label: v_p_seasonal
#| fig-height: 2.5
#| fig-aspect: .7
#| out-width: 100%

# Bring CRP/DCP determinants and compute seasonal proportions per grid cell
df <- prepare_period_df(a_det$crp, a_det$dcp, c("pwnt","psmr","ptot")) |>
  filter(is.finite(ptot), is.finite(pwnt), is.finite(psmr), area > 0, ptot > 0) |>
  mutate(
    winter_prop = pwnt / ptot,
    summer_prop = psmr / ptot
  )

# Area-weighted mean proportions by period
agg_pct <- df |>
  group_by(period_lbl) |>
  summarise(
    Winter = weighted.mean(winter_prop, w = area, na.rm = TRUE),
    Summer = weighted.mean(summer_prop, w = area, na.rm = TRUE),
    .groups = "drop"
  ) |>
  tidyr::pivot_longer(cols = c(Winter, Summer),
                      names_to = "season", values_to = "prop") |>
  mutate(
    season = factor(season, levels = c("Winter","Summer")),
    period_lbl = factor(period_lbl, levels = c(CRP_LBL, DCP_LBL))
  )

# Compute a compact upper limit: +15% headroom, cap at 100%, round up to nearest 5%
y_max <- min(1, max(agg_pct$prop, na.rm = TRUE) * 1.15)
y_max <- 0.05 * ceiling(y_max / 0.05)
if (!is.finite(y_max) || y_max <= 0) y_max <- 0.05

fill_vals <- role_palette$season

# Side-by-side bars with percent y-axis
p_seasonal <- ggplot(agg_pct, aes(x = season, y = prop, fill = season)) +
  geom_col(width = 0.6, color = "gray20") +
  facet_wrap(~ period_lbl, ncol = 2) +
  scale_fill_manual(values = fill_vals) +
  scale_y_continuous(
    labels = scales::percent_format(accuracy = 1),
    limits = c(0, y_max),
    expand = expansion(mult = c(0, 0.03))
  ) +
  labs(
    title = "Seasonal Percentage of Total Annual Precipitation",
    subtitle = "Used in Aridity Threshold Calculation",
    x = NULL, y = "Share of Annual Precipitation (%)"
  ) +
  theme(strip.text = element_text(face = "bold"), legend.position = "none")

p_seasonal

```

### Aridity

Plot — Aridity boundary (ptot vs athr). Each point is a grid cell. Points below the dashed 1:1 line (ptot < athr) are dry; those above are humid. Compare period densities relative to the boundary.


```{r}
#| label: v_a_dry
#| fig-height: 5
#| out-width: 100%

# Prepare data and calculate max athr
aridity_data <- prepare_period_df(a_det$crp, a_det$dcp, c("athr", "ptot")) |>
  filter(is.finite(athr), athr > 0, area > 0) %>%
  mutate(
    period = pretty_period_lbl(period),
    dry_status = factor(if_else(ptot < athr, "Dry", "Humid"),
                        levels = c("Dry", "Humid"))
  )
max_athr <- max(aridity_data$athr, na.rm = TRUE)

# Plot aridity threshold
p_aridity <- aridity_data |>
  ggplot(aes(x = athr, weight = area, fill = dry_status)) +
  geom_histogram(binwidth = 50, boundary = 0, color = "white") +
  scale_fill_manual(values = role_palette$dryness) +
  scale_x_continuous(breaks = seq(0, ceiling(max_athr), by = 200)) +
  labs(
    title = "Aridity Threshold",
    subtitle = "Identifies Dry Climates (when greater than Total Precipitation)",
    x = "Aridity Threshold (mm)",
    y = "Weighted Count (by Area)",
    fill = "Climate Status"
  ) +
  facet_wrap(~ period, ncol = 1) +
  theme(legend.position = "top")

p_aridity
```

```{r}
#| label: v_a_scatter
#| fig-height: 7
#| out-width: 100%

pt_size <- 0.25
pt_alpha <- 0.15

ar_scatter <- prepare_period_df(a_det$crp, a_det$dcp, c("athr", "ptot", "tmax")) %>%
  filter(is.finite(ptot), is.finite(athr), ptot > 0, athr > 0, tmax > 10, area > 0) %>%
  mutate(
    period = pretty_period_lbl(period),
    dry_status = factor(if_else(ptot < athr, "Dry", "Humid"),
                        levels = c("Dry", "Humid"))
  )

ggplot(ar_scatter, aes(x = athr, y = ptot, color = dry_status)) +
  # keep the boundary line out of the legend + give it a neutral color
  geom_abline(slope = 1, intercept = 0, linetype = "dashed",
              color = "grey50", inherit.aes = FALSE) +
  geom_point(size = pt_size, alpha = pt_alpha) +
  scale_color_manual(
    name   = NULL,
    values = c(Dry = role_palette$dryness[["Dry"]],
               Humid = role_palette$dryness[["Humid"]]),
    breaks = c("Dry","Humid")
  ) +
  # 2) start x-axis at 1 (drops any athr < 1) and
  # 3) show whole-number labels (no scientific notation)
  scale_x_log10(
    limits = c(1, NA),
    labels = scales::label_number(accuracy = 1, big.mark = ",")
  ) +
  scale_y_log10(
    labels = scales::label_number(accuracy = 1, big.mark = ",")
  ) +
  coord_equal() +
  labs(
    title = "Total Precipitation vs Aridity Threshold",
    subtitle = "Dashed line denotes the dry boundary (ptot = athr)",
    x = "Aridity Threshold (mm, log10)",
    y = "Total Precipitation (mm, log10)"
  ) +
  # 1) make legend swatches clearly visible (not faded)
  guides(color = guide_legend(override.aes = list(alpha = 1, size = 3))) +
  facet_wrap(~ period_lbl, ncol = 2) +
  theme(legend.position = "top")
  
```

## Save Results
```{r}
#| label: save_results
write_parquet(a_det$crp, file.path(odir, "crp_determinants.parquet"))
write_parquet(a_det$dcp, file.path(odir, "dcp_determinants.parquet"))
```
