---
title: "Global Climate Classification Changes"
subtitle: "Climate Reference (1961–1990) vs Recent Climate (2015–2024)"
abstract: "Identify changes using Köppen and Trewartha classification systems."
author: "Rick Forest"
date: today
format:
  html:
    code-fold: true            # reader option to view or not view code
    code-fold-default: hide    # code cells start collapsed (optional)
    code-tools: true           # adds "Show All Code / Hide All Code" buttons
    toc: true
    number-sections: true
  pdf: default
execute:
  warning: false
  message: false
  echo: true
  output: true
---

## Setup

### Packages

```{r setup}
suppressPackageStartupMessages({
  library(tidyverse)
  library(ecmwfr)
  library(tidync)
  library(data.table)
  library(sf)
  library(arrow)
  library(units)

  library(gt)
  #library(gtExtras)
  #library(htmltools)

  library(RColorBrewer)
  library(viridis)
  library(scales) # for cut_number
  library(rnaturalearth)
})
```

### Parameters

```{r parameters}

classification_method <- "Köppen–Trewartha"

# climate reference period (crp)
crp_beg <- 1961
crp_end <- 1990
crp_lbl <- paste("CRP ", crp_beg, "-", crp_end, sep = "")

# decadal climate normals (dcn)
dcn_beg <- 2015
dcn_end <- 2024
dcn_lbl <- paste("DCN ", dcn_beg, "-", dcn_end, sep = "")


rbn_prj <- "ESRI:54030"
```

### Paths

```{r paths}
dir_download <- "data/1-download/"
dir_transform <- "data/2-transform/"
```

### Dictionaries

```{r ktc-dictionary}

ktc_zone <- c(
    "Ar",
    "Aw",
    "BW",
    "BS",
    "Cs",
    "Cf",
    "Do",
    "Dc",
    "E",
    "Ft",
    "Fi"
  )

ktc_label <- c(
    "Ar: Tropical Wet",
    "Aw: Tropical Wet & Dry",
    "BW: Desert",
    "BS: Steppe",
    "Cs: Mediterranean",
    "Cf: Subtropical",
    "Do: Temperate Oceanic",
    "Dc: Temperate Continental",
    "E:  Sub-Arctic",
    "Ft: Polar Tundra",
    "Fi: Polar Ice Cap"
  )

ktc_dict  <- data.frame(zone= ktc_zone, label=ktc_label)
```

```{r kcc-dictionary}

kcc_zone <- c(
    "Af",
    "Am",
    "As",
    "Aw",
    "BSh",
    "BSk",
    "BWh",
    "BWk",
    "Cfa",
    "Cfb",
    "Cfc",
    "Csa",
    "Csb",
    "Csc",
    "Cwa",
    "Cwb",
    "Cwc",
    "Dfa",
    "Dfb",
    "Dfc",
    "Dfd",
    "Dsa",
    "Dsb",
    "Dsc",
    "Dsd",
    "Dwa",
    "Dwb",
    "Dwc",
    "Dwd",
    "ET",
    "EF"
  )

kcc_label <- c(
    "Af: Tropical (Rainforest)",
    "Am: Tropical (Monsoon)",
    "As: Tropical (Savanna, Dry S)",
    "Aw: Tropical (Savanna, Dry W)",
    "BSh: Dry (Semi-arid steppe Hot)",
    "BSk: Dry (Semi-arid steppe Cold)",
    "BWh: Dry (Arid desert Hot)",
    "BWk: Dry (Arid desert Cold)",
    "Cfa: Temperate (No dry season Hot S)",
    "Cfb: Temperate (No dry season Warm S)",
    "Cfc: Temperate (No dry season Cold S)",
    "Csa: Temperate (Dry summer Hot S)",
    "Csb: Temperate (Dry summer Warm S)",
    "Csc: Temperate (Dry summer Cold S)",
    "Cwa: Temperate (Dry winter Hot S)",
    "Cwb: Temperate (Dry winter Warm S)",
    "Cwc: Temperate (Dry winter Cold S)",
    "Dfa: Continental (No dry season Hot S)",
    "Dfb: Continental (No dry season Warm S)",
    "Dfc: Continental (No dry season Cold S)",
    "Dfd: Continental (No dry season Very cold S)",
    "Dsa: Continental (Dry winter Hot S)",
    "Dsb: Continental (Dry winter Warm S)",
    "Dsc: Continental (Dry winter Cold S)",
    "Dsd: Continental (Dry winter Very cold S)",
    "Dwa: Continental (Dry summer Hot S)",
    "Dwb: Continental (Dry summer Warm S)",
    "Dwc: Continental (Dry summer Cold S)",
    "Dwd: Continental (Dry summer Very cold S)",
    "ET: Polar (Tundra)",
    "EF: Polar Ice Cap"
  )

kcc_dict  <- data.frame(zone= kcc_zone, label=kcc_label)

```

### Palettes

```{r ktc palette}

 ktc_color <- c(
    "#040ff0",
    "#0474fb",
    "#ffe505",
    "#ff9005",
    "#0eff00",
    "#2db83d",
    "#b64dff",
    "#7c00ff",
    "#017783",
    "#72b8ce",
    "#b4dfe3"
  )

ktc_palette  <- data.frame(zone= ktc_zone, color=ktc_color)

```

```{r kcc palette}

 kcc_color <- c(
    "#0000fe",
    "#0277ff",
    "#7abaec",
    "#45a9f9",
    "#f5a300",
    "#ffdb63",
    "#fe0100",
    "#fe9695", 
    "#c7ff4f",
    "#66ff33",
    "#32c702",
    "#ffff00",
    "#c6c701",
    "#969600",
    "#96ff96",
    "#63c764",
    "#329633",
    "#03ffff",
    "#37c7ff",
    "#007e7d",
    "#00455e",
    "#ff00fe",
    "#c600c7",
    "#863295",
    "#966495",
    "#abb1ff",
    "#5a77db",
    "#4c51b5",
    "#320087",
    "#b2b2b2",
    "#686868"
  )

kcc_palette  <- data.frame(zone= kcc_zone, color=kcc_color)

```

## Download data

### Download ERA5 climate data

```{r ERA5-download-ERA5-climate-data}

fn_download_ERA5_climate_data <- function(pth, var, beg, end) {
  yrs <- beg:end
  mths <- 1:12
  if (var == "t") {
    variable_name <- "2m_temperature"
  } else if (var == "p") {
    variable_name <- "total_precipitation"
  }
  request <- list(
    product_type = "monthly_averaged_reanalysis",
    dataset_short_name = "reanalysis-era5-single-levels-monthly-means",
    variable = variable_name,
    year = yrs,
    month = mths,
    time = c("00:00"),
    format = "netcdf",
    download_format = "unarchived",
    target = pth
  )
  # Climate Data Store key
  options(keyring_backend = "file") # for ecmwfr to use file instead of mac keychain
  wf_get_key() # keyring password required 
  file <- wf_request(request = request, transfer = TRUE, path = ".")
}

fn_download_climate_period_data <- function(dir, typ, var, beg, end) {
  fil <- paste(typ, "-era5-", var, ".nc", sep = "")
  pth <- paste(dir, fil, sep = "")
  if (!file.exists(pth)) {
    fn_download_ERA5_climate_data(pth, var, beg, end) 
  }
}

dir <- dir_download

fn_download_climate_period_data(dir, "crp", "t", crp_beg, crp_end)
fn_download_climate_period_data(dir, "crp", "p", crp_beg, crp_end)

fn_download_climate_period_data(dir, "dcn", "t", dcn_beg, dcn_end)
fn_download_climate_period_data(dir, "dcn", "p", dcn_beg, dcn_end)

```

### Download ERA5 land-sea mask

```{r download-land-sea-mask} 

fn_download_ERA5_land_sea_mask <- function(pth) {
  request <- list(
    product_type = "reanalysis",
    dataset_short_name = "reanalysis-era5-single-levels",
    variable = "land_sea_mask",
    year = "2000", # any date works; the mask is invariant
    month = "01",
    day = "01",
    time = "00:00",
    area = "90/-179.75/-90/180", # N/W/S/E (global)
    grid = "0.25/0.25", # ERA5 native grid spacing
    format = "netcdf",
    download_format = "unarchived",
    target = pth
  )
  # Climate Data Store key
  options(keyring_backend = "file") # for ecmwfr to use file instead of mac keychain
  wf_get_key() # keyring password required 
  file <- wf_request(request = request, transfer = TRUE, path = ".")
}

dir <- dir_download
fil <- "land-sea-mask.nc"
pth <- paste(dir, fil, sep = "")
if (!file.exists(pth))  {
  fn_download_ERA5_land_sea_mask(pth)
}
```

### Download global coastline

```{r download-global-coastline}

fn_download_global_coastline <- function() {
  coastline_sf <-
    ne_download(
      scale = 10,
      type = "coastline",
      category = "physical",
      returnclass = "sf"
    )
    return(coastline_sf)
}

dir <- dir_download
fil <- "global-coastline.gpkg"
pth <- paste(dir, fil, sep = "")

if (!file.exists(pth))  {
  coastline_sf <- fn_download_global_coastline()
  st_write(coastline_sf, pth, delete_dsn = TRUE, quiet = TRUE)
}

```

## Transform data

### Transform land sea mask

```{r transform-land-sea-mask}

fn_transform_land_sea_mask <- function(dir_download) {

  fil <- "land-sea-mask.nc"
  pth <- paste(dir_download, fil, sep = "")
  nc <- tidync(pth)

  df <- nc |>
    hyper_tibble() |>
    as_tibble() |> # tidyverse data frame
    rename(lon = longitude, lat = latitude) |>
    mutate(
      lon = as.numeric(lon),
      lat = as.numeric(lat),
      lsm = as.integer(round(lsm))
    ) |>
    select(lon, lat, lsm) |> # drop all other columns
    arrange(lon, lat)
  return(df)
}

dir <- dir_transform
fil <- "land-sea-mask.parquet"
pth <- paste(dir, fil, sep = "")

if (!file.exists(pth))  {
  land_sea_mask_df <- fn_transform_land_sea_mask(dir_download)
  write_parquet(land_sea_mask_df, pth)
} else {
  land_sea_mask_df <- read_parquet(pth)
}

```

### Transform global coastline

```{r transform-global-coastline}

fn_transform_global_coastline <- function(dir_download) {
  # Create a single coastline multiline, making sure geometries are valid
  # (eliminates extra horizontal line on map)

  fil <- "global-coastline.gpkg"
  pth <- paste(dir_download, fil, sep = "")
  coastline_sf <- st_read(pth, quiet = TRUE)

  coastline_sf <- coastline_sf |>
    select(geom) |>
    mutate(id = "world") |>
    group_by(id) |>
    st_make_valid() |>
    summarise()
  return(coastline_sf)
}

dir <- dir_transform
fil <- "global-coastline.gpkg"
pth <- paste(dir, fil, sep = "")

if (!file.exists(pth))  {
  coastline_sf <- fn_transform_global_coastline(dir_download)
  st_write(coastline_sf, pth, delete_dsn = TRUE, quiet = TRUE)
} else {
  coastline_sf <- st_read(pth, quiet=TRUE)
}

```

### Read NetCDF climate data

```{r read-nc-chunk}

fn_read_nc_chunk <- function(dt, var, vnm, start, stop, ncs_df) {

  setnames(dt, old = "longitude", new = "lon")
  setnames(dt, old = "latitude", new = "lat")
  setnames(dt, old = vnm, new = var)

  dt[, lon := as.numeric(lon)]
  dt[, lat := as.numeric(lat)]
  dt[lon > 180, lon := lon - 360]

  dt[, mth := as.integer(substr(valid_time, 6, 7))]
  setDT(dt)

  # add row before grouping
  cv <- vnm
  if (var == "t") {
    avgcv <- round(mean(dt$t) - 273.15, 3)
    maxcv <- round(max(dt$t) - 273.15, 3)
    mincv <- round(min(dt$t) - 273.15, 3)
    maxdate <- max(dt$valid_time)
    mindate <- min(dt$valid_time)
  } else {
    avgcv <- round(mean(dt$p) * 1000, 3)
    maxcv <- round(max(dt$p) * 1000, 3)
    mincv <- round(min(dt$p) * 1000, 3)
    maxdate <- substr(max(dt$valid_time), 1, 10)
    mindate <- substr(min(dt$valid_time), 1, 10)
  }
  maxlon <- max(dt$lon)
  minlon <- min(dt$lon)
  maxlat <- max(dt$lat)
  minlat <- min(dt$lat)
  datapts <- nrow(dt)

  new_row <- list(
    cv = cv,
    avgcv   = avgcv,
    maxcv   = maxcv,
    mincv   = mincv,
    maxdate = maxdate,
    mindate = mindate,
    maxlon  = maxlon,
    minlon  = minlon,
    maxlat  = maxlat,
    minlat  = minlat,
    datapts = datapts
  )

  ncs_df <- rbind(ncs_df, new_row)

  if (var == "t") {
    drop <- setdiff(names(dt), c("lon", "lat", "mth", "t"))
    dt[, (drop) := NULL]
    dt <- dt[, .(t = sum(t, na.rm = TRUE)), by = .(lon, lat, mth)]
  } else {
    drop <- setdiff(names(dt), c("lon", "lat", "mth", "p"))
    dt[, (drop) := NULL]
    dt <- dt[, .(p = sum(p, na.rm = TRUE)), by = .(lon, lat, mth)]
  }
  setDT(dt)
  gc()

  return(list(dt, ncs_df))
}

```

```{r read-nc}
fn_read_nc <- function(pth, var, vnm, ncs_df) {
  nc <- tidync(pth)

  dims <- nc |> activate(vnm) |> hyper_dims()
  ntime <- dims |> filter(name == "valid_time") |> pull(length)
  chunk <- 60L

  print(paste("chunk size ", chunk))

  for (start in seq(1L, ntime, by = chunk)) {
    chnk_beg <- Sys.time()
    stop <- min(start + chunk - 1L, ntime)

    chunk_dt <- nc |>
      hyper_filter(valid_time = index %in% start:stop) |>
      hyper_tibble(select_var = vnm) |>
      as.data.table()


    result <- fn_read_nc_chunk(chunk_dt, var, vnm, start, stop, ncs_df)
    chunk_grp_dt <- result[[1]]
    ncs_df <- result[[2]]

    if (start == 1L) {
      dt <- chunk_grp_dt
    } else {
      dt <- rbindlist(list(dt, chunk_grp_dt))
      # group
      if (var == "t") {
        dt <- dt[, .(t = sum(t, na.rm = TRUE)), by = .(lon, lat, mth)]
      } else {
        dt <- dt[, .(p = sum(p, na.rm = TRUE)), by = .(lon, lat, mth)]
      }
    }
    chnk_end <- Sys.time()
    print(paste("chunk time: ", (chnk_end - chnk_beg)))
  }

  return(list(dt, ncs_df))
}

```

```{r get-t-data}

fn_get_t_data <- function(dir, typ, ncs_df) {

  var <- "t"
  vnm <- "t2m"
  fil <- paste(typ, "-era5-", var, ".nc", sep = "")
  pth <- paste(dir, fil, sep = "")
  result <- fn_read_nc(pth, var, vnm, ncs_df)
  t_dt <- result[[1]]
  ncs_df <- result[[2]]

  setorder(t_dt, lon, lat, mth)
  setDT(t_dt)

  # convert from sum to avg
  yrs <- ifelse(typ == "crp", 30, 10)
  t_dt[, t := t / yrs]

  # convert from kelvin to celsius
  t_dt[, t := t - 273.15]

  return(list(t_dt, ncs_df))
}
```

```{r get-p-data}

fn_get_p_data <- function(dir, typ, ncs_df) {

  ndy <- data.table(
    mth = 1:12,
    ndays = as.integer(c(31, 28, 31, 30, 31, 30, 31, 31, 30, 31, 30, 31))
  )

  var <- "p"
  vnm <- "tp"
  fil <- paste(typ, "-era5-", var, ".nc", sep = "")
  pth <- paste(dir, fil, sep = "")
  result <- fn_read_nc(pth, var, vnm, ncs_df)
  p_dt <- result[[1]]
  ncs_df <- result[[2]]

  setorder(p_dt, lon, lat, mth)
  setDT(p_dt)

  # convert from sum to avg
  yrs <- ifelse(typ == "crp", 30, 10)
  p_dt[, p := p / yrs]

  # add number of days using lookup table
  p_dt[ndy, ndy := i.ndays, on = "mth"]
  p_dt[, p := p * ndy * 1000]
  p_dt[, ndy := NULL]

  return(list(p_dt, ncs_df))

}
```

### Transform NetCDF climate data

```{r transform-t-data, typ}

fn_transform_t_data <- function(t_dt) {

  # pivot and calculate stats

  t_wide_dt <- t_dt |>
    pivot_wider(
      names_from = c(mth),
      values_from = c("t"),
      names_prefix = "t",
      names_sep = "" # no underscore in new col names
    )

  tcols <- paste0("t", 1:12)
  t_stats_dt <- t_wide_dt |>
    mutate(
      tavg = (t1 + t2 + t3 + t4 + t5 + t6 + t7 + t8 + t9 + t10 + t11 + t12) /
        12,
      tmax = pmax(t1, t2, t3, t4, t5, t6, t7, t8, t9, t10, t11, t12),
      tmin = pmin(t1, t2, t3, t4, t5, t6, t7, t8, t9, t10, t11, t12),
      totm = (t1 + t2 + t3 + t10 + t11 + t12) / 6,
      tats = (t4 + t5 + t6 + t7 + t8 + t9) / 6,
      tg10 = rowSums(across(all_of(tcols), ~ .x >= 10))
    )
  t_stats_df <- t_stats_dt |>
    select(-t1, -t2, -t3, -t4, -t5, -t6, -t7, -t8, -t9, -t10, -t11, -t12)

  return(t_stats_df)
}

```

```{r transform-p-data}

fn_transform_p_data <- function(p_dt) {

  # pivot and calculate stats

  p_wide_dt <- p_dt |>
    pivot_wider(
      names_from = c(mth),
      values_from = c("p"),
      names_prefix = "p",
      names_sep = "" # no underscore in new col names
    )
  
  pcols <- paste0("p", 1:12)
  p_stats_dt <- p_wide_dt |>
    mutate(
      ptot = (p1 + p2 + p3 + p4 + p5 + p6 + p7 + p8 + p9 + p10 + p11 + p12),
      pmax = pmax(p1, p2, p3, p4, p5, p6, p7, p8, p9, p10, p11, p12),
      pmin = pmin(p1, p2, p3, p4, p5, p6, p7, p8, p9, p10, p11, p12),
      pavg = ptot / 12,
      potm = (p1 + p2 + p3 + p10 + p11 + p12),
      pats = (p4 + p5 + p6 + p7 + p8 + p9),
      pmno = pmin(p10, p11, p12, p1, p2, p3),
      pmxo = pmax(p10, p11, p12, p1, p2, p3),
      pmna = pmin(p4, p5, p6, p7, p8, p9),
      pmxa = pmax(p4, p5, p6, p7, p8, p9),
      pl60 = rowSums(across(all_of(pcols), ~ .x < 60))
    )
  p_stats_df <- p_stats_dt |>
    select(-p1, -p2, -p3, -p4, -p5, -p6, -p7, -p8, -p9, -p10, -p11, -p12)

  return(p_stats_df)

}

```

### Calculate grid cell area

```{r calc-grid-area}

fn_calc_area <- function(df) {
  earth_radius_km <- 6371
  lat_grid_degrees <- 0.25
  lon_grid_degrees <- 0.25
  area_df <- df |>
    mutate(
      lat_radians = lat * pi / 180,
      lat_grid_radians = lat_grid_degrees * pi / 180,
      lon_grid_radians = lon_grid_degrees * pi / 180,
      # cell area in km^2
      area = (earth_radius_km^2 *
        lon_grid_radians *
        (sin(lat_radians + lat_grid_radians / 2) -
          sin(lat_radians - lat_grid_radians / 2))) /
        1e6
    ) |>
    select(-lat_radians, -lat_grid_radians, -lon_grid_radians) |>
    select(lat, lon, area, lsm, everything())
  return(area_df)

}
```

### Save climate data

```{r save-climate-data}

fn_create_ncs_df <- function() {
  ncs_df <- data.frame(
    cv      = character(),
    avgcv   = numeric(),
    maxcv   = numeric(),
    mincv   = numeric(),
    maxdate = character(),
    mindate = character(),
    maxlon  = numeric(),
    minlon  = numeric(),
    maxlat  = numeric(),
    minlat  = numeric(),
    datapts = numeric()
  )
  return(ncs_df)
}

fn_save_climate_data <- function (dir_read, dir_save, typ, ncs_df) {
 
  result <- fn_get_t_data(dir_read, typ, ncs_df)
  t_dt <- result[[1]]
  ncs_df <- result[[2]]
  pth <- paste(dir_save, typ, "-monthly-data-t.parquet", sep = "")
  write_parquet(t_dt, pth)

  result <- fn_get_p_data(dir_read, typ, ncs_df)
  p_dt <- result[[1]]
  ncs_df <- result[[2]]
  pth <- paste(dir_save, typ, "-monthly-data-p.parquet", sep = "")
  write_parquet(p_dt, pth)

  pth <- paste(dir_save, typ, "-nc-summary.csv", sep = "")
  write_csv(ncs_df, pth)

  t_stats_df <- fn_transform_t_data(t_dt)
  pth <- paste(dir_save, typ, "-average-stats-t.parquet", sep = "")
  write_parquet(t_stats_df, pth)

  p_stats_df <- fn_transform_p_data(p_dt)
  pth <- paste(dir_save, typ, "-average-stats-p.parquet", sep = "")
  write_parquet(p_stats_df, pth)

  j1_df <- inner_join(t_stats_df, p_stats_df, by = c("lon", "lat"))
  j2_df <- inner_join(j1_df, land_sea_mask_df, by = c("lon", "lat"))
  cn_df <- fn_calc_area(j2_df)
  pth <- paste(dir_save, typ, "-climate-normals.parquet", sep = "")
  write_parquet(cn_df, pth)

}

dir_read <- dir_download
dir_save <- dir_transform

typ <- "dcn"
pth <- paste(dir_save, typ, "-climate-normals.parquet", sep = "")
# only need to read, transform, and save if file doesn't exist
if (!file.exists(pth)) {
  ncs_df <- fn_create_ncs_df()
  fn_save_climate_data(dir_read, dir_save, typ, ncs_df) 
}

typ <- "crp"
pth <- paste(dir_save, typ, "-climate-normals.parquet", sep = "")
# only need to read, transform, and save if file doesn't exist
if (!file.exists(pth)) {
  ncs_df <- fn_create_ncs_df()
  fn_save_climate_data(dir_read, dir_save, typ, ncs_df)  
}

```

### Read climate period data

```{r read-crp-saved-data}

typ <- "crp"

pth <- paste(dir_transform, typ, "-nc-summary.csv", sep = "")
crp_nc_summary_df <- read_csv(pth, show_col_types = FALSE) 

pth <- paste(dir_transform, typ, "-monthly-data-t.parquet", sep = "")
crp_t_df <- read_parquet(pth) 
pth <- paste(dir_transform, typ, "-monthly-data-p.parquet", sep = "")
crp_p_df <- read_parquet(pth) 
pth <- paste(dir_transform, typ, "-average-stats-t.parquet", sep = "")
crp_t_stats_df <- read_parquet(pth) 
pth <- paste(dir_transform, typ, "-average-stats-p.parquet", sep = "")
crp_p_stats_df <- read_parquet(pth) 
pth <- paste(dir_transform, typ, "-climate-normals.parquet", sep = "")
crp_cn_df <- read_parquet(pth) 

typ <- "dcn"

pth <- paste(dir_transform, typ, "-nc-summary.csv", sep = "")
crp_nc_summary_df <- read_csv(pth, show_col_types = FALSE) 

pth <- paste(dir_transform, typ, "-monthly-data-t.parquet", sep = "")
dcn_t_df <- read_parquet(pth) 
pth <- paste(dir_transform, typ, "-monthly-data-p.parquet", sep = "")
dcn_p_df <- read_parquet(pth) 
pth <- paste(dir_transform, typ, "-average-stats-t.parquet", sep = "")
dcn_t_stats_df <- read_parquet(pth) 
pth <- paste(dir_transform, typ, "-average-stats-p.parquet", sep = "")
dcn_p_stats_df <- read_parquet(pth) 
pth <- paste(dir_transform, typ, "-climate-normals.parquet", sep = "")
dcn_cn_df <- read_parquet(pth) 

```

### Summary of netcdf data

```{r netcdf-summary-data}
# fn_create_nc_summary <- function (df1, df2) {

#   #summary_df <- df |> 
  

#   summary_tbl <- tibble(
#     Metric = c(
#       "Climate variable",
#       "Climate variable avg",
#       "Climate variable max",
#       "Climate variable min",
#       "End date",
#       "Start date",
#       "Max longitude",
#       "Min longitude",
#       "Max latitude",
#       "Min latitude",
#       "Data points"
#     ),
#     CRP_t = c(
#       nrow(df),
#       max(df$lon, na.rm = TRUE),
#       min(df$lon, na.rm = TRUE),
#       max(df$lat, na.rm = TRUE),
#       min(df$lat, na.rm = TRUE),
#       mean(df$t1, na.rm = TRUE),
#       max(df$t1, na.rm = TRUE),
#       min(df$t1, na.rm = TRUE)
#     ),
#     DCN_t = c(
#       nrow(df),
#       max(df$lon, na.rm = TRUE),
#       min(df$lon, na.rm = TRUE),
#       max(df$lat, na.rm = TRUE),
#       min(df$lat, na.rm = TRUE),
#       mean(df$t2, na.rm = TRUE),
#       max(df$t2, na.rm = TRUE),
#       min(df$t2, na.rm = TRUE)
#     ),
#     CRP_p = c(
#       nrow(df),
#       max(df$lon, na.rm = TRUE),
#       min(df$lon, na.rm = TRUE),
#       max(df$lat, na.rm = TRUE),
#       min(df$lat, na.rm = TRUE),
#       mean(df$p1, na.rm = TRUE),
#       max(df$p1, na.rm = TRUE),
#       min(df$p1, na.rm = TRUE)
#     ),
#     DCN_p = c(
#       nrow(df),
#       max(df$lon, na.rm = TRUE),
#       min(df$lon, na.rm = TRUE),
#       max(df$lat, na.rm = TRUE),
#       min(df$lat, na.rm = TRUE),
#       mean(df$p2, na.rm = TRUE),
#       max(df$p2, na.rm = TRUE),
#       min(df$p2, na.rm = TRUE)
#     )
#   )
 
#   summary_tbl <- summary_df %>%
#     gt() %>%
#       fmt_number(
#         columns = c(CRP_t, DCN_t, CRP_p , DCN_p),
#         decimals = 0,
#         use_seps = TRUE,
#           rows = Metric == "Total rows"
#       ) %>%
#       fmt_number(
#         columns = c(CRP_t, DCN_t, CRP_p , DCN_p),
#         decimals = 2,
#         rows = Metric %in% c(
#           "Max longitude", "Min longitude",
#           "Max latitude", "Min latitude"
#         )
#       ) %>%
#       fmt_number(
#         columns = c(CRP_t, DCN_t, CRP_p , DCN_p),
#         decimals = 3,
#         rows = Metric %in% c(
#           "Climate Variable Avg",
#           "Climate Variable Max",
#           "Climate Variable Min"
#         )
#       ) %>%
#       tab_spanner(
#         label = "Temperature (°C)",
#         columns = c(CRP_t,DCN_t)
#       ) %>%
#       tab_spanner(
#         label = "Precipitation (mm)",
#         columns = c(CRP_p,DCN_p)
#       ) %>%
#       cols_label(
#         CRP_t = lbl1,
#         DCN_t = lbl2,
#         CRP_p = lbl1,
#         DCN_p = lbl2
#       ) %>%
#       tab_header(
#         title = title
#         # ,
#         # subtitle = subtitle
#       ) %>%
#       tab_style(
#         style = list(
#           cell_fill(color = "lightyellow")
#         ),
#         locations = cells_body(
#           rows = Metric %in% c(
#             "Climate Variable Avg"
#           )
#         )
#       )

#   return(summary_tbl)

# }

# tbl_crp_nc_gt <- fn_create_nc_summary(crp_nc_summary_df)
# tbl_dcn_nc_gt <- fn_create_nc_summary(dcn_nc_summary_df)

```

## Explore data

```{r summary-table-data}

joined_df <- crp_t_df %>%
  select(lat, lon, mth, t) %>%
  rename(t1 = t) %>%
  inner_join(dcn_t_df %>% select(lat, lon, mth, t) %>% rename(t2 = t),
             by = c("lat", "lon", "mth")) %>%
  inner_join(crp_p_df %>% select(lat, lon, mth, p) %>% rename(p1 = p),
             by = c("lat", "lon", "mth")) %>%
  inner_join(dcn_p_df %>% select(lat, lon, mth, p) %>% rename(p2 = p),
             by = c("lat", "lon", "mth"))

fn_create_summary_table_4x <- function (df, title, subtitle, lbl1, lbl2) {

  summary_df <- tibble(
    Metric = c(
      "Data points",
      "Max longitude",
      "Min longitude",
      "Max latitude",
      "Min latitude",
      "Climate Variable Avg",
      "Climate Variable Max",
      "Climate Variable Min"
    ),
    CRP_t = c(
      nrow(df),
      max(df$lon, na.rm = TRUE),
      min(df$lon, na.rm = TRUE),
      max(df$lat, na.rm = TRUE),
      min(df$lat, na.rm = TRUE),
      mean(df$t1, na.rm = TRUE),
      max(df$t1, na.rm = TRUE),
      min(df$t1, na.rm = TRUE)
    ),
    DCN_t = c(
      nrow(df),
      max(df$lon, na.rm = TRUE),
      min(df$lon, na.rm = TRUE),
      max(df$lat, na.rm = TRUE),
      min(df$lat, na.rm = TRUE),
      mean(df$t2, na.rm = TRUE),
      max(df$t2, na.rm = TRUE),
      min(df$t2, na.rm = TRUE)
    ),
    CRP_p = c(
      nrow(df),
      max(df$lon, na.rm = TRUE),
      min(df$lon, na.rm = TRUE),
      max(df$lat, na.rm = TRUE),
      min(df$lat, na.rm = TRUE),
      mean(df$p1, na.rm = TRUE),
      max(df$p1, na.rm = TRUE),
      min(df$p1, na.rm = TRUE)
    ),
    DCN_p = c(
      nrow(df),
      max(df$lon, na.rm = TRUE),
      min(df$lon, na.rm = TRUE),
      max(df$lat, na.rm = TRUE),
      min(df$lat, na.rm = TRUE),
      mean(df$p2, na.rm = TRUE),
      max(df$p2, na.rm = TRUE),
      min(df$p2, na.rm = TRUE)
    )
  )
 
  summary_tbl <- summary_df %>%
    gt() %>%
      fmt_number(
        columns = c(CRP_t, DCN_t, CRP_p , DCN_p),
        decimals = 0,
        use_seps = TRUE,
          rows = Metric == "Total rows"
      ) %>%
      fmt_number(
        columns = c(CRP_t, DCN_t, CRP_p , DCN_p),
        decimals = 2,
        rows = Metric %in% c(
          "Max longitude", "Min longitude",
          "Max latitude", "Min latitude"
        )
      ) %>%
      fmt_number(
        columns = c(CRP_t, DCN_t, CRP_p , DCN_p),
        decimals = 3,
        rows = Metric %in% c(
          "Climate Variable Avg",
          "Climate Variable Max",
          "Climate Variable Min"
        )
      ) %>%
      tab_spanner(
        label = "Temperature (°C)",
        columns = c(CRP_t,DCN_t)
      ) %>%
      tab_spanner(
        label = "Precipitation (mm)",
        columns = c(CRP_p,DCN_p)
      ) %>%
      cols_label(
        CRP_t = lbl1,
        DCN_t = lbl2,
        CRP_p = lbl1,
        DCN_p = lbl2
      ) %>%
      tab_header(
        title = title
        # ,
        # subtitle = subtitle
      ) %>%
      tab_style(
        style = list(
          cell_fill(color = "lightyellow")
        ),
        locations = cells_body(
          rows = Metric %in% c(
            "Climate Variable Avg"
          )
        )
      )

  return(summary_tbl)
}

title <- "Summary of Input Climate Variables"
lbl1 <- crp_lbl
lbl2 <- dcn_lbl
tbl1_gt <- fn_create_summary_table_4x(joined_df, title, subtitle, lbl1, lbl2)
tbl1_gt

```

## Exploratory Plots

### Climate Variable

```{r expl-cv-plot}

fn_explore_cv_plot <- function(df, typ) {
  if (typ == "t") {
    cv_df <- df |> 
       select(lon, lat, tavg) |> 
       rename(cv = tavg)
    cv_title <- paste("Average Temperature ", crp_lbl)
    all_colors <- rev(brewer.pal(11, "RdBu"))
    cv_colors <- all_colors[c(4, 5, 7, 8, 9)]    
    cv_units <- "°C"
  } else {
    cv_df <- df |> 
       select(lon, lat, ptot) |> 
       rename(cv = ptot)  
    cv_title <- paste("Average Precipitation ", crp_lbl)
    cv_colors <- brewer.pal(5, "Blues") 
    cv_units <- "mm"
  }
  qs <- quantile(cv_df$cv, probs = seq(0, 1, length.out = 6), na.rm = TRUE)  
  labs <- sprintf("[%.1f, %.1f]", qs[-6], qs[-1]) 
  cv_df <- cv_df %>%
    mutate(p_bin = cut(cv, breaks = qs, include.lowest = TRUE, labels = labs))

  cv_plt <- ggplot(cv_df, aes(lon, lat, fill = p_bin)) +
    geom_raster(alpha = 0.85) +
    geom_sf(data = coastline_sf,
      color = "black", linewidth = 0.15,
      inherit.aes= FALSE, show.legend = FALSE) +
    scale_fill_manual(
      values = cv_colors,
      name = cv_units, 
      guide = guide_legend(
        reverse = TRUE,       # high value bin at top
        override.aes = list(
          alpha = 0.85,        # legend bins semi-transparent
          colour = "black",   # legend bins outline
          size = 0.4          # legend bins outline line width
        )   
      )   
    ) +
    coord_sf(expand=FALSE) +
    labs(
      title = cv_title,
      x = "",
      y = "") +
    theme_minimal() +
    # make legend bins boxes
    theme(legend.key = element_rect(colour = "black", fill = NA)) 
 return (cv_plt)
}
p1t_exp_t_crp <- fn_explore_cv_plot(crp_cn_df,"t")
p1t_exp_p_crp <- fn_explore_cv_plot(crp_cn_df,"p")
p1t_exp_t_crp
p1t_exp_p_crp 
```

### Climate variable changes

```{r expl-cv-plot-chg}

fn_cv_chg_vis <- function(df1, df2, typ) {
  if (typ == "t") {
    cv_df1 <- df1 |> select(lon, lat, tavg) |> rename(cv = tavg)
    cv_df2 <- df2 |> select(lon, lat, tavg) |> rename(cv = tavg)
    cv_title <- paste("Temperature Changes ", dcn_lbl, " vs ", crp_lbl)
    all_colors <- rev(brewer.pal(11, "RdBu"))
    cv_colors <- all_colors[c(4, 5, 7, 8, 9)]  
    cv_units <- "°C"
  } else {
    cv_df1 <- df1 |> select(lon, lat, ptot) |> rename(cv = ptot)
    cv_df2 <- df2 |> select(lon, lat, ptot) |> rename(cv = ptot)
    cv_title <- paste("Precipitation Changes", dcn_lbl, " vs ", crp_lbl)
    all_colors <- brewer.pal(11, "PRGn")
    cv_colors <- all_colors[3:7]
    cv_units <- "mm"
  }
  join_df <- cv_df1 |>
    select(lat, lon, cv) |>
    rename(cv1 = cv) |>
    inner_join(cv_df2  |> 
    select(lat, lon, cv)  |> 
    rename(cv2 = cv), by = c("lat", "lon")) |> 
    mutate(cvdiff = cv2 - cv1) |> 
    select(lon, lat, cvdiff)
  qs <- quantile(join_df$cvdiff, 
                probs = seq(0, 1, length.out = 6),
                na.rm = TRUE)
  labs <- sprintf("[%.1f, %.1f]", qs[-length(qs)], qs[-1])
  plot_df <- join_df |>
    mutate(
      p_bin = cut(
        cvdiff,
        breaks = qs,
        include.lowest = TRUE,
        right = TRUE,
        labels = labs
      )
    )  
  cv_plt <- ggplot(plot_df, aes(x = lon, y = lat, fill = p_bin)) +
    geom_raster(alpha = 0.85) +
    geom_sf(data = coastline_sf,
      color = "black", linewidth = 0.15,
      inherit.aes= FALSE, show.legend = FALSE) +
    scale_fill_manual(
      values = cv_colors,
      name = cv_units, 
      na.translate = FALSE,
      guide = guide_legend(
        reverse = TRUE,       # high value bin at top
        override.aes = list(
          alpha = 0.85,        # legend bins semi-transparent
          colour = "black",   # legend bins outline
          size = 0.4          # legend bins outline line width
        )   
      )   
    ) +
    coord_sf(expand=FALSE) +
    labs(
      title = cv_title,
      x = "",
      y = "") +
    theme_minimal() +
    # make legend bins boxes
    theme(legend.key = element_rect(colour = "black", fill = NA)) 
}
p1t_exp_t_chg <- fn_cv_chg_vis(crp_cn_df, dcn_cn_df, "t")
p1t_exp_p_chg <- fn_cv_chg_vis(crp_cn_df, dcn_cn_df, "p")
p1t_exp_t_chg
p1t_exp_p_chg 
```

## Classification

### KTC classification function

```{r classify_ktc}
fn_classify_ktc <- function(df) {

# Step 1: Add seasonal precip stats (Summer/Winter) based on hemisphere
#         Summer: high sun 6 months (SH: Oct–Mar, NH: Apr–Sep)
#         Winter: low  sun 6 months (NH: Oct–Mar, SH: Apr–Sep)
  df <- df |>
    mutate(
      pST = case_when(lat >= 0 ~ pats, TRUE ~ potm),
      pWT = case_when(lat >= 0 ~ potm, TRUE ~ pats),
      pMnST = case_when(lat >= 0 ~ pmna, TRUE ~ pmno),
      pMxST = case_when(lat >= 0 ~ pmxa, TRUE ~ pmxo),
      pMnWT = case_when(lat >= 0 ~ pmno, TRUE ~ pmna),
      pMxWT = case_when(lat >= 0 ~ pmxo, TRUE ~ pmxa)
    )

# Step 2: Calculate Aridity threshold
  df <- df |>
    mutate(
      athr = case_when(
        ptot != 0 ~ (10 * (tavg - 10)) + (3 * (pST / ptot) * 100),
        TRUE ~ 10 * (tavg - 10)
      )
    )

# Step 3: Assign Primary Group
  df <- df |>
    mutate(
      z1 = case_when(
        tmax < 10 ~ "F",
        ptot < (athr * 2) ~ "B",
        tmin >= 18 ~ "A",
        tg10 >= 8 ~ "C",
        tg10 >= 4 ~ "D",
        tg10 >= 1 ~ "E",
        TRUE ~ "X"
      ),
      zn = ""
    )

# Step 4: Assign secondary group
  df <- df |>
    mutate(
      z2 = case_when(
        z1 == "A" & pl60 <= 2 ~ "r",
        z1 == "A" ~ "w",
        z1 == "B" & ptot < athr ~ "W",
        z1 == "B" ~ "S",
        z1 == "C" & pWT >= pST * 3 & pmin < 30 & ptot < 890 ~ "s",
        z1 == "C" ~ "f",
        z1 == "D" & tmin > 0 ~ "o",
        z1 == "D" ~ "c",
        z1 == "F" & tmax > 0 ~ "t",
        z1 == "F" ~ "i",
        TRUE ~ ""
      ),
      zn = paste(z1, z2, sep = "")
    )
  return(df)
}
```

### Classify climate periods using KTC

```{r ktc-crp-zones}

ktc_crp_cn_df <- fn_classify_ktc(crp_cn_df)
ktc_crp_cn_df <- ktc_crp_cn_df |> select(lon, lat, zn, area, lsm)
ktc_crp_cn_sf <- st_as_sf(ktc_crp_cn_df, coords = c("lon", "lat"), crs = 4326)

ktc_d_df <- fn_classify_ktc(dcn_cn_df)
ktc_d_df <- ktc_d_df |> select(lon, lat, zn, area, lsm)
ktc_d_sf <- st_as_sf(ktc_d_df, coords = c("lon", "lat"), crs = 4326)

# combine areas by crp and dcn

ktc_cd_sf <- ktc_crp_cn_sf
ktc_cd_sf <- ktc_cd_sf |> rename(czn = zn)
ktc_cd_sf$dzn <- ktc_d_sf$zn

ktc_grp_geom_sf <- ktc_cd_sf |>
  filter(lsm==1) |>
  summarise(
    geometry = st_combine(geometry),
    mkm = sum(area),
    .by = c(czn, dzn)
  ) |>
  mutate(cz1 = substr(czn, 1, 1)) |>
  mutate(dz1 = substr(dzn, 1, 1)) |>
  mutate(pct = round(mkm / sum(mkm) * 100, 3)) |>
  select(czn, cz1, dzn, dz1, mkm, pct, geometry) |>
  arrange(czn, dzn)
ktc_grp_geom_sf <- st_set_crs(ktc_grp_geom_sf, 4326)

```


### KCC classification function

 ```{r kcc-seasonal-stats}
fn_classify_kcc <- function(df) {

# Step 1: Add seasonal precip stats (Summer/Winter) based on hemisphere
#         Summer: high sun 6 months (SH: Oct–Mar, NH: Apr–Sep)
#         Winter: low  sun 6 months (NH: Oct–Mar, SH: Apr–Sep)
  df <- df |>
    mutate(
      pSK = case_when(lat >= 0 ~ pats, TRUE ~ potm),
      pWK = case_when(lat >= 0 ~ potm, TRUE ~ pats),
      pMnSK = case_when(lat >= 0 ~ pmna, TRUE ~ pmno),
      pMxSK = case_when(lat >= 0 ~ pmxa, TRUE ~ pmxo),
      pMnWK = case_when(lat >= 0 ~ pmno, TRUE ~ pmna),
      pMxWK = case_when(lat >= 0 ~ pmxo, TRUE ~ pmxa)
    )

# Step 2: Calculate Precipitation threshold
  df <- df |>
    mutate(
      pthr = case_when(
        pSK / ptot >= 0.7 ~ tavg * 20 + 280,
         pSK / ptot >= 0.3 ~ tavg * 20 + 140,
         TRUE ~ tavg * 20
      )
    )

# Step 3: Assign Primary Group
  df <- df |>
    mutate(
      z1 = case_when(
        tmax < 10 ~ "E",
        ptot < pthr ~ "B",
        tmin > 18 ~ "A",
        tmax > 10 & tmin >= -3 ~ "C",
        tmax > 10 & tmin <= -3 ~ "D",
        TRUE ~ ""
      ),
      z2 = ""
      # ,
      # k3 = "",
    )

# Step 4: Assign 2nd letter
  df <- df |>
    mutate(
      z2 = case_when(
        z1 == "A" & pmin >= 60 ~ "f",
        z1 == "A" & pmin < 60 & pmin >= (100 - ptot / 25) ~ "m",
        z1 == "A" & pmin < (100 - ptot / 25) & pWK > pSK ~ "w",
        z1 == "A" ~ "s",

        z1 == "B" & ptot <= (pthr * 0.5) ~ "W",
        z1 == "B" ~ "S",

        z1 == "C" & pMnSK < 40 & pMxWK >= pMnSK * 3 ~ "s",
        z1 == "C" & pMxSK >= pMnWK * 10 ~ "w",
        z1 == "C" ~ "f",

        z1 == "D" & pMnSK < 40 & pMxWK >= pMnSK * 3 ~ "s",
        z1 == "D" & pMxSK >= pMnWK * 10 ~ "w",
        z1 == "D" ~ "f",

        z1 == "E" & tmax > 0 ~ "T",
        z1 == "E" & tmax <= 0 ~ "F",

        TRUE ~ ""
      ),
    zn = paste(z1, z2, sep = "")    
    )

# Step 5: Assign 3rd letter
  df <- df |> 
    mutate(
      z3 = case_when(
        z1 == "B" & tavg >= 18 ~ "h",
        z1 == "B" ~ "k",

        z1 == "C" & tmax >= 22 ~ "a",
        z1 == "C" & tg10 >= 4 ~ "b",
        z1 == "C" ~ "c",

        z1 == "D" & tmax >= 22 ~ "a",
        z1 == "D" & tg10 >= 4 ~ "b",
        z1 == "D" & tmin < -38 ~ "d",
        z1 == "D" ~ "c",

        TRUE ~ ""
      ),
    zn = paste(z1, z2, z3,sep = "")  
    )


}

# Alternate threshold definition:
#   # Summer in Koppen is defined as warmer 6 month period of Oct-Mar or Apr-Sep
#   # Winter in Koppen is defined as cooler 6 month period of Oct-Mar or Apr-Sep
#   dt <- dt |>
#     mutate(
#       pSK = case_when(
#         tAprSep >= tOctMar ~ pats,
#         TRUE ~ potm
#       ),
#       pWK = case_when(
#         tAprSep < tOctMar ~ pats,
#         TRUE ~ potm
#       ),
#       pMnSK = case_when(
#         tAprSep >= tOctMar ~ pmin(p4, p5, p6, p7, p8, p9),
#         TRUE ~ pmin(p1, p2, p3, p10, p11, p12),
#       ),
#       pMxSK = case_when(
#         tAprSep >= tOctMar ~ pmax(p4, p5, p6, p7, p8, p9),
#         TRUE ~ pmax(p1, p2, p3, p10, p11, p12),
#       ),
#       pMnWK = case_when(
#         tAprSep >= tOctMar ~ pmin(p1, p2, p3, p10, p11, p12),
#         TRUE ~ pmin(p4, p5, p6, p7, p8, p9),
#       ),
#       pMxWK = case_when(
#         tAprSep >= tOctMar ~ pmax(p1, p2, p3, p10, p11, p12),
#         TRUE ~ pmax(p4, p5, p6, p7, p8, p9),
#       ),
#     )
#   return(dt)
# }

 ```

### Classify climate periods using KCC

```{r kcc-crp-zones}

kcc_crp_cn_df <- fn_classify_kcc(crp_cn_df)
kcc_crp_cn_df <- kcc_crp_cn_df |> select(lon, lat, zn, area, lsm)
kcc_crp_cn_sf <- st_as_sf(kcc_crp_cn_df, coords = c("lon", "lat"), crs = 4326)

kcc_dcn_cn_df <- fn_classify_kcc(dcn_cn_df)
kcc_dcn_cn_df <- kcc_dcn_cn_df |> select(lon, lat, zn, area, lsm)
kcc_dcn_cn_sf <- st_as_sf(kcc_dcn_cn_df, coords = c("lon", "lat"), crs = 4326)

# combine areas by crp and dcn

kcc_cd_sf <- kcc_crp_cn_sf
kcc_cd_sf <- kcc_cd_sf |> rename(czn = zn)
kcc_cd_sf$dzn <- kcc_dcn_cn_sf$zn

kcc_grp_geom_sf <- kcc_cd_sf |>
  filter(lsm==1) |>
  summarise(
    geometry = st_combine(geometry),
    mkm = sum(area),
    .by = c(czn, dzn)
  ) |>
  mutate(cz1 = substr(czn, 1, 1)) |>
  mutate(dz1 = substr(dzn, 1, 1)) |>
  mutate(pct = round(mkm / sum(mkm) * 100, 3)) |>
  select(czn, cz1, dzn, dz1, mkm, pct, geometry) |>
  arrange(czn, dzn)
kcc_grp_geom_sf <- st_set_crs(kcc_grp_geom_sf, 4326)


```


## Summary

### KTC summary

```{r build-ktc-summary}

# Build CRP and DCN aggregates from grouped land geometries
ktc_crp_df <- ktc_grp_geom_sf |>
  st_drop_geometry() |>
  select(czn, cz1, mkm, pct) |>
  summarize(mkm = sum(mkm), pct = sum(pct), .by = c(czn, cz1)) |>
  mutate(period = "crp") |>
  rename(zone = czn, z1 = cz1) |>
  arrange(zone)

ktc_dcn_df <- ktc_grp_geom_sf |>
  st_drop_geometry() |>
  select(dzn, dz1, mkm, pct) |>
  summarize(mkm = sum(mkm), pct = sum(pct), .by = c(dzn, dz1)) |>
  mutate(period = "dcn") |>
  rename(zone = dzn, z1 = dz1) |>
  arrange(zone)

# Side-by-side comparison + deltas
ktc_diff_df <- ktc_crp_df |>
  select(zone, mkm, pct) |>
  rename(cmkm = mkm, cpct = pct)

ktc_diff_df$dmkm <- ktc_dcn_df$mkm
ktc_diff_df$dpct <- ktc_dcn_df$pct

ktc_diff_df <- ktc_diff_df |>
  mutate(
    xmkm = dmkm - cmkm,
    xpct = round(dpct - cpct, 2)
  ) |>
  select(zone, cmkm, dmkm, xmkm, cpct, dpct, xpct)

ktc_diff_df <- ktc_diff_df |>
  left_join(ktc_dict, by = "zone") |>
  select(label, cmkm, dmkm, xmkm, cpct, dpct, xpct)

ktc_diff_gt <- gt(ktc_diff_df)

```

```{r display-ktc-summary}

ktc_tbl1 <- ktc_diff_gt |>
  tab_header(
    title = "Summary of Köppen-Trewartha Climate Zone Changes",
    #subtitle = paste(crp_lbl, "vs", dcn_lbl)
  ) |>
  tab_spanner(
    label = "Area (Mkm²)",
    columns = c(cmkm, dmkm, xmkm)
  ) |>
  tab_spanner(
    label = "Earth Land Area %",
    columns = c(cpct, dpct, xpct)
  ) |>
  cols_label(
    cmkm = "CRP",
    dmkm = "DCN",
    xmkm = "Chg",
    cpct = "CRP",
    dpct = "DCN",
    xpct = "Chg"
  ) |>
  fmt_number(columns = c(cmkm, dmkm, xmkm), decimals = 1) |>
  fmt_number(columns = c(cpct, dpct, xpct), decimals = 2) |>
  grand_summary_rows(
    columns = c(cmkm, dmkm, xmkm, cpct, dpct, xpct),
    fns = list("Total" = ~ sum(.)),
    fmt = list(~ fmt_number(.x, decimals = 2))
  ) |>
  tab_footnote(
    footnote = paste("CRP: Climate Reference Period ", crp_beg, "-", crp_end, sep = ""),
    locations = cells_column_labels(columns = c(cmkm, cpct))
  ) |>
  tab_footnote(
    footnote = paste("DCN: Decadal Climate Normals ", dcn_beg, "-", dcn_end, sep = ""),
    locations = cells_column_labels(columns = c(dmkm, dpct))
  )

ktc_tbl1 
```

### KCC summary

```{r build-kcc-summary}

# Build CRP and DCN aggregates from grouped land geometries
kcc_crp_df <- kcc_grp_geom_sf |>
  st_drop_geometry() |>
  select(czn, cz1, mkm, pct) |>
  summarize(mkm = sum(mkm), pct = sum(pct), .by = c(czn, cz1)) |>
  mutate(period = "crp") |>
  rename(zone = czn, z1 = cz1) |>
  arrange(zone)

kcc_dcn_df <- kcc_grp_geom_sf |>
  st_drop_geometry() |>
  select(dzn, dz1, mkm, pct) |>
  summarize(mkm = sum(mkm), pct = sum(pct), .by = c(dzn, dz1)) |>
  mutate(period = "dcn") |>
  rename(zone = dzn, z1 = dz1) |>
  arrange(zone)

# Side-by-side comparison + deltas
kcc_diff_df <- kcc_crp_df |>
  select(zone, mkm, pct) |>
  rename(cmkm = mkm, cpct = pct)

kcc_diff_df <- kcc_diff_df |>
  inner_join(kcc_dcn_df, by = "zone") |>
  rename(dmkm = mkm, dpct = pct)

# kcc_diff_df$dmkm <- kcc_dcn_df$mkm
# kcc_diff_df$dpct <- kcc_dcn_df$pct

kcc_diff_df <- kcc_diff_df |>
  mutate(
    xmkm = dmkm - cmkm,
    xpct = round(dpct - cpct, 2)
  ) |>
  select(zone, cmkm, dmkm, xmkm, cpct, dpct, xpct)

kcc_diff_df <- kcc_diff_df |>
  inner_join(kcc_dict, by = "zone") |>
  select(label, cmkm, dmkm, xmkm, cpct, dpct, xpct)

kcc_diff_gt <- gt(kcc_diff_df)

```

```{r display-kcc-summary}

kcc_tbl1 <- kcc_diff_gt |>
  tab_header(
    title = "Summary of Köppen Climate Zone Changes",
    #subtitle = paste(crp_lbl, "vs", dcn_lbl)
  ) |>
  tab_spanner(
    label = "Area (Mkm²)",
    columns = c(cmkm, dmkm, xmkm)
  ) |>
  tab_spanner(
    label = "Earth Land Area %",
    columns = c(cpct, dpct, xpct)
  ) |>
  cols_label(
    cmkm = "CRP",
    dmkm = "DCN",
    xmkm = "Chg",
    cpct = "CRP",
    dpct = "DCN",
    xpct = "Chg"
  ) |>
  fmt_number(columns = c(cmkm, dmkm, xmkm), decimals = 1) |>
  fmt_number(columns = c(cpct, dpct, xpct), decimals = 2) |>
  grand_summary_rows(
    columns = c(cmkm, dmkm, xmkm, cpct, dpct, xpct),
    fns = list("Total" = ~ sum(.)),
    fmt = list(~ fmt_number(.x, decimals = 2))
  ) |>
  tab_footnote(
    footnote = paste("CRP: Climate Reference Period ", crp_beg, "-", crp_end, sep = ""),
    locations = cells_column_labels(columns = c(cmkm, cpct))
  ) |>
  tab_footnote(
    footnote = paste("DCN: Decadal Climate Normals ", dcn_beg, "-", dcn_end, sep = ""),
    locations = cells_column_labels(columns = c(dmkm, dpct))
  )

kcc_tbl1 

```



## Visualization

### Map 1

```{r map-1}

# plot_data <- ktc_grp_geom_sf |>
#   select(czn) |>
#   rename(zone = czn)
# p1 <- ggplot(plot_data) +
#   geom_sf(aes(color = zone), size = 0.1) +
#   scale_color_manual(
#     values = setNames(ktc_palette$color, ktc_palette$zone),
#     labels = setNames(ktc_dict$label, ktc_dict$zone)
#   ) +
#   geom_sf(
#     data = coastline_sf,
#     color = "darkgrey",
#     linewidth = 0.1
#   ) +
#   coord_sf(crs = rbn_prj) +
#   labs(
#     title = "Köppen-Trewartha Climate Classification",
#     subtitle = crp_lbl,
#     color = "Zones"
#   ) +
#   guides(
#     color = guide_legend(
#       override.aes = list(
#         size = 3 # legend point size only
#       )
#     )
#   )
# p1

```

### Map 2

```{r map-2}

# plot_data <- ktc_grp_geom_sf |>
#   select(dzn) |>
#   rename(zone = dzn)
# p2 <- ggplot(plot_data) +
#   geom_sf(aes(color = zone), size = 0.1) +
#   scale_color_manual(
#     values = setNames(ktc_palette$color, ktc_palette$zone),
#     labels = setNames(ktc_dict$label, ktc_dict$zone)
#   ) +
#   geom_sf(
#     data = coastline_sf,
#     color = "darkgrey",
#     linewidth = 0.1
#   ) +
#   coord_sf(crs = rbn_prj) +
#   labs(
#     title = "Köppen-Trewartha Climate Classification",
#     subtitle = dcn_lbl,
#     color = "Zones"
#   ) +
#   guides(
#     color = guide_legend(
#       override.aes = list(
#         size = 3 # legend point size only
#       )
#     )
#   )
# #p2

```

### Map 3

```{r map-3}

# plot_z1 <- "B"
# plot_data <- ktc_grp_geom_sf |>
#   filter(cz1 != dz1) |>
#   filter(dz1 == plot_z1) |>
#   select(czn) |>
#   rename(zone = czn)
# p3 <- ggplot(plot_data) +
#   geom_sf(aes(color = zone), size = 0.1) +
#   scale_color_manual(
#     values = setNames(ktc_palette$color, ktc_palette$zone),
#     labels = setNames(ktc_dict$label, ktc_dict$zone)
#   ) +
#   geom_sf(
#     data = coastline_sf,
#     color = "darkgrey",
#     linewidth = 0.1
#   ) +
#   coord_sf(crs = rbn_prj) +
#   labs(
#     title = "Köppen-Trewartha Climate Classification",
#     subtitle = paste0(
#       "New Type ",
#       plot_z1,
#       " Zones in ",
#       dcn_lbl
#     ),
#     color = "Source Zones"
#   ) +
#   guides(
#     color = guide_legend(
#       override.aes = list(
#         size = 4 # legend point size only
#       )
#     )
#   )
# #p3

```

### Map 4

```{r map-4}

# plot_zn <- "Cs"
# plot_data <- ktc_grp_geom_sf |>
#   filter(dzn == plot_zn) |>
#   rename(zone = dzn)

# p4 <- ggplot(plot_data) +
#   geom_sf(aes(color = zone), size = 0.1) +
#   scale_color_manual(
#     values = setNames(ktc_palette$color, ktc_palette$zone),
#     labels = setNames(ktc_dict$label, ktc_dict$zone)
#   ) +
#   geom_sf(
#     data = coastline_sf,
#     color = "darkgrey",
#     linewidth = 0.1
#   ) +
#   coord_sf(crs = rbn_prj) +
#   labs(
#     title = "Köppen-Trewartha Climate Classification",
#     subtitle = dcn_lbl,
#     color = "Zones"
#   ) +
#   guides(
#     color = guide_legend(
#       override.aes = list(
#         size = 4 # legend point size only
#       )
#     )
#   )
# #p4

```

### Map 5

```{r map-5}

# plot_data <- kcc_grp_geom_sf |>
#   select(czn) |>
#   rename(zone = czn)

# p5 <- ggplot(plot_data) +
#   geom_sf(aes(color = zone), size = 0.1) +
#   scale_color_manual(
#     values = setNames(kcc_palette$color, kcc_palette$zone),
#     labels = setNames(kcc_dict$label, kcc_dict$zone)
#   ) +
#   geom_sf(
#     data = coastline_sf,
#     color = "darkgrey",
#     linewidth = 0.1
#   ) +
#   coord_sf(crs = rbn_prj) +
#   labs(
#     title = "Köppen Climate Classification",
#     subtitle = crp_lbl,
#     color = "Zones"
#   ) +
#   guides(
#     color = guide_legend(
#       override.aes = list(
#         size = 3 # legend point size only
#       )
#     )
#   )
# p5

```

### Map 6

```{r map-6}

# plot_data <- kcc_grp_geom_sf |>
#   select(dzn) |>
#   rename(zone = dzn)
# p6 <- ggplot(plot_data) +
#   geom_sf(aes(color = zone), size = 0.1) +
#   scale_color_manual(
#     values = setNames(kcc_palette$color, kcc_palette$zone),
#     labels = setNames(kcc_dict$label, kcc_dict$zone)
#   ) +
#   geom_sf(
#     data = coastline_sf,
#     color = "darkgrey",
#     linewidth = 0.1
#   ) +
#   coord_sf(crs = rbn_prj) +
#   labs(
#     title = "Köppen Climate Classification",
#     subtitle = dcn_lbl,
#     color = "Zones"
#   ) +
#   guides(
#     color = guide_legend(
#       override.aes = list(
#         size = 3 # legend point size only
#       )
#     )
#   )
# #p6

```

### Map 7

```{r map-7}

# plot_z1 <- "B"
# plot_data <- kcc_grp_geom_sf |>
#   filter(cz1 != dz1) |>
#   filter(dz1 == plot_z1) |>
#   select(czn) |>
#   rename(zone = czn)
# p7 <- ggplot(plot_data) +
#   geom_sf(aes(color = zone), size = 0.1) +
#   scale_color_manual(
#     values = setNames(kcc_palette$color, kcc_palette$zone),
#     labels = setNames(kcc_dict$label, kcc_dict$zone)
#   ) +
#   geom_sf(
#     data = coastline_sf,
#     color = "darkgrey",
#     linewidth = 0.1
#   ) +
#   coord_sf(crs = rbn_prj) +
#   labs(
#     title = "Köppen Climate Classification",
#     subtitle = paste0(
#       "New Type ",
#       plot_z1,
#       " Zones in ",
#       dcn_lbl
#     ),
#     color = "Source Zones"
#   ) +
#   guides(
#     color = guide_legend(
#       override.aes = list(
#         size = 4 # legend point size only
#       )
#     )
#   )
# #p7

```

### Map 8

```{r map-8}

# plot_zn <- "Cs"
# plot_data <- kcc_grp_geom_sf |>
#   filter(dzn == plot_zn) |>
#   rename(zone = dzn)

# p8 <- ggplot(plot_data) +
#   geom_sf(aes(color = zone), size = 0.1) +
#   scale_color_manual(
#     values = setNames(kcc_palette$color, kcc_palette$zone),
#     labels = setNames(kcc_dict$label, kcc_dict$zone)
#   ) +
#   geom_sf(
#     data = coastline_sf,
#     color = "darkgrey",
#     linewidth = 0.1
#   ) +
#   coord_sf(crs = rbn_prj) +
#   labs(
#     title = "Köppen Climate Classification",
#     subtitle = dcn_lbl,
#     color = "Zones"
#   ) +
#   guides(
#     color = guide_legend(
#       override.aes = list(
#         size = 4 # legend point size only
#       )
#     )
#   )
# #p8

```
