---
title: "Global Climate Zone Changes"
subtitle: "Climate Reference (1961–1990)\n vs Recent Climate (2015–2024)"
abstract: "Identify changes using Köppen and Trewartha classification systems."
author: "Rick Forest"
date: today
format:
  html:
    code-fold: true            # reader option to view or not view code
    code-fold-default: hide    # code cells start collapsed (optional)
    code-tools: true           # adds "Show All Code / Hide All Code" buttons
    toc: true
    number-sections: true
  pdf: default
execute:
  warning: false
  message: false
---

## Setup

### Packages

```{r setup}
suppressPackageStartupMessages({
  library(tidyverse)
  library(ecmwfr)
  library(tidync)
  library(data.table)
  library(sf)
  library(arrow)
  library(units)
  library(gt)
  library(viridis)
  library(scales)
  library(rnaturalearth)
})
```

### Parameters

```{r parameters}

classification_method <- "Köppen–Trewartha"

# climate reference period (crp)
crp_beg <- 1961
crp_end <- 1990
crp_lbl <- paste("CRP ", crp_beg, "-", crp_end, sep = "")

# decadal climate normals (dcn)
dcn_beg <- 2015
dcn_end <- 2024
dcn_lbl <- paste("DCN ", dcn_beg, "-", dcn_end, sep = "")

ndy <- data.table(
  mth = 1:12,
  ndays = as.integer(c(31, 28, 31, 30, 31, 30, 31, 31, 30, 31, 30, 31))
)

rbn_prj <- "ESRI:54030"
```

### Paths

```{r paths}
dir_download <- "data/1-download/"
dir_transform <- "data/2-transform/"
```

### Dictionaries

```{r dictionaries}

  zone <- c(
    "Ar",
    "Aw",
    "BW",
    "BS",
    "Cs",
    "Cf",
    "Do",
    "Dc",
    "E",
    "Ft",
    "Fi"
  )

  label <- c(
    "Ar: Tropical Wet",
    "Aw: Tropical Wet & Dry",
    "BW: Desert",
    "BS: Steppe",
    "Cs: Mediterranean",
    "Cf: Subtropical",
    "Do: Temperate Oceanic",
    "Dc: Temperate Continental",
    "E:  Sub-Arctic",
    "Ft: Polar Tundra",
    "Fi: Polar Ice Cap"
  )

ktc_dict  <- data.frame(zone= zone, label=label)
```

### Palettes

```{r palettes}

 ktc_color <- c(
    "#040ff0",
    "#0474fb",
    "#ffe505",
    "#ff9005",
    "#0eff00",
    "#2db83d",
    "#b64dff",
    "#7c00ff",
    "#017783",
    "#72b8ce",
    "#b4dfe3"
  )

ktc_palette  <- data.frame(zone= zone, color=ktc_color)

```

## Download data

### Download ERA5 climate data

```{r ERA5-download-ERA5-climate-data}

fn_download_ERA5_climate_data <- function(pth, var, beg, end) {
  yrs <- beg:end
  mths <- 1:12
  if (var == "t") {
    variable_name <- "2m_temperature"
  } else if (var == "p") {
    variable_name <- "total_precipitation"
  }
  request <- list(
    product_type = "monthly_averaged_reanalysis",
    dataset_short_name = "reanalysis-era5-single-levels-monthly-means",
    variable = variable_name,
    year = yrs,
    month = mths,
    time = c("00:00"),
    format = "netcdf",
    download_format = "unarchived",
    target = pth
  )
  # Climate Data Store key
  options(keyring_backend = "file") # for ecmwfr to use file instead of mac keychain
  wf_get_key() # keyring password required 
  file <- wf_request(request = request, transfer = TRUE, path = ".")
}

fn_download_climate_period_data <- function(dir, typ, var, beg, end) {
  fil <- paste(typ, "-", beg, "-", end, "-", var, ".nc", sep = "")
  pth <- paste(dir, fil, sep = "")
  if (!file.exists(pth)) {
    fn_download_ERA5_climate_data(pth, var, beg, end) 
  }
}

dir <- dir_download

fn_download_climate_period_data(dir, "crp", "t", crp_beg, crp_end)
fn_download_climate_period_data(dir, "crp", "p", crp_beg, crp_end)

fn_download_climate_period_data(dir, "dcn", "t", dcn_beg, dcn_end)
fn_download_climate_period_data(dir, "dcn", "p", dcn_beg, dcn_end)

```

### Download ERA5 land-sea mask

```{r download-land-sea-mask} 

fn_download_ERA5_land_sea_mask <- function(pth) {
  request <- list(
    product_type = "reanalysis",
    dataset_short_name = "reanalysis-era5-single-levels",
    variable = "land_sea_mask",
    year = "2000", # any date works; the mask is invariant
    month = "01",
    day = "01",
    time = "00:00",
    area = "90/-179.75/-90/180", # N/W/S/E (global)
    grid = "0.25/0.25", # ERA5 native grid spacing
    format = "netcdf",
    download_format = "unarchived",
    target = pth
  )
  # Climate Data Store key
  options(keyring_backend = "file") # for ecmwfr to use file instead of mac keychain
  wf_get_key() # keyring password required 
  file <- wf_request(request = request, transfer = TRUE, path = ".")
}

dir <- dir_download
fil <- "land-sea-mask.nc"
pth <- paste(dir, fil, sep = "")
if (!file.exists(pth))  {
  fn_download_ERA5_land_sea_mask(pth)
}
```

### Download global coastline

```{r download-global-coastline}

fn_download_global_coastline <- function() {
  coastline_sf <-
    ne_download(
      scale = 10,
      type = "coastline",
      category = "physical",
      returnclass = "sf"
    )
    return(coastline_sf)
}

dir <- dir_download
fil <- "global-coastline.gpkg"
pth <- paste(dir, fil, sep = "")

if (!file.exists(pth))  {
  coastline_sf <- fn_download_global_coastline()
  st_write(coastline_sf, pth, delete_dsn = TRUE, quiet = TRUE)
}

```

## Transform data

### Transform land sea mask

```{r transform-land-sea-mask}

fn_transform_land_sea_mask <- function(dir_download) {

  fil <- "land-sea-mask.nc"
  pth <- paste(dir_download, fil, sep = "")
  nc <- tidync(pth)

  df <- nc |>
    hyper_tibble() |>
    as_tibble() |> # tidyverse data frame
    rename(lon = longitude, lat = latitude) |>
    mutate(
      lon = as.numeric(lon),
      lat = as.numeric(lat),
      lsm = as.integer(round(lsm))
    ) |>
    select(lon, lat, lsm) |> # drop all other columns
    arrange(lon, lat)
  return(df)
}

dir <- dir_transform
fil <- "land-sea-mask.parquet"
pth <- paste(dir, fil, sep = "")

if (!file.exists(pth))  {
  land_sea_mask_df <- fn_transform_land_sea_mask(dir_download)
  write_parquet(land_sea_mask_df, pth)
} else {
  land_sea_mask_df <- read_parquet(pth)
}

```

### Transform global coastline

```{r transform-global-coastline}

fn_transform_global_coastline <- function(dir_download) {
  # Create a single coastline multiline, making sure geometries are valid
  # (eliminates extra horizontal line on map)

  fil <- "global-coastline.gpkg"
  pth <- paste(dir_download, fil, sep = "")
  coastline_sf <- st_read(pth)

  coastline_sf <- coastline_sf |>
    select(geom) |>
    mutate(id = "world") |>
    group_by(id) |>
    st_make_valid() |>
    summarise()
  return(coastline_sf)
}

dir <- dir_transform
fil <- "global-coastline.gpkg"
pth <- paste(dir, fil, sep = "")

if (!file.exists(pth))  {
  coastline_sf <- fn_transform_global_coastline(dir_download)
  st_write(coastline_sf, pth, delete_dsn = TRUE, quiet = TRUE)
} else {
  coastline_sf <- st_read(pth)
}

```

### Read NetCDF climate data

```{r read-nc-chunk}

fn_read_nc_chunk <- function(dt, var, vnm, start, stop) {
  setnames(dt, old = "longitude", new = "lon")
  setnames(dt, old = "latitude", new = "lat")
  setnames(dt, old = vnm, new = var)

  dt[, lon := as.numeric(lon)]
  dt[, lat := as.numeric(lat)]
  dt[lon > 180, lon := lon - 360]

  dt[, mth := as.integer(substr(valid_time, 6, 7))]
  setDT(dt)

  if (var == "t") {
    drop <- setdiff(names(dt), c("lon", "lat", "mth", "t"))
    dt[, (drop) := NULL]
    dt <- dt[, .(t = sum(t, na.rm = TRUE)), by = .(lon, lat, mth)]
  } else {
    drop <- setdiff(names(dt), c("lon", "lat", "mth", "p"))
    dt[, (drop) := NULL]
    dt <- dt[, .(p = sum(p, na.rm = TRUE)), by = .(lon, lat, mth)]
  }
  setDT(dt)
  gc()

  return(dt)
}

```

```{r read-nc}
fn_read_nc <- function(pth, var, vnm) {
  nc <- tidync(pth)

  dims <- nc |> activate(vnm) |> hyper_dims()
  ntime <- dims |> filter(name == "valid_time") |> pull(length)
  chunk <- 60L

  print(paste("chunk size ", chunk))

  for (start in seq(1L, ntime, by = chunk)) {
    chnk_beg <- Sys.time()
    stop <- min(start + chunk - 1L, ntime)

    chunk_dt <- nc |>
      hyper_filter(valid_time = index %in% start:stop) |>
      hyper_tibble(select_var = vnm) |>
      as.data.table()

    chunk_grp_dt <- fn_read_nc_chunk(chunk_dt, var, vnm, start, stop)

    fn2_pt2_beg = Sys.time()
    if (start == 1L) {
      dt <- chunk_grp_dt
    } else {
      dt <- rbindlist(list(dt, chunk_grp_dt))
      # group
      if (var == "t") {
        dt <- dt[, .(t = sum(t, na.rm = TRUE)), by = .(lon, lat, mth)]
      } else {
        dt <- dt[, .(p = sum(p, na.rm = TRUE)), by = .(lon, lat, mth)]
      }
    }
    chnk_end <- Sys.time()
    print(paste("chunk time: ", (chnk_end - chnk_beg)))
  }

  return(dt)
}

```

```{r}
fn_get_t_data <- function(dir, typ, beg, end) {
  var <- "t"
  vnm <- "t2m"
  fil <- paste(typ, "-", beg, "-", end, "-", var, ".nc", sep = "")
  pth <- paste(dir, fil, sep = "")
  t_dt <- fn_read_nc(pth, var, vnm)
  return(t_dt)
}
```

```{r}
fn_get_p_data <- function(dir, typ, beg, end, ndy) {
  var <- "p"
  vnm <- "tp"
  fil <- paste(typ, "-", beg, "-", end, "-", var, ".nc", sep = "")
  pth <- paste(dir, fil, sep = "")
  p_dt <- fn_read_nc(pth, var, vnm)
  return(p_dt)
}
```

### Transform NetCDF climate data

```{r transform-t-data, typ}

fn_transform_t_data <- function(t_dt, typ) {
  yrs <- ifelse(typ == "crp", 30, 10)

  # sort, convert to period avg, convert units

  setorder(t_dt, lon, lat, mth)
  setDT(t_dt)
  # convert from sum to avg
  t_dt[, t := t / yrs]
  # convert from kelvin to celsius
  t_dt[, t := t - 273.15]

  # pivot

  t_wide_dt <- t_dt |>
    pivot_wider(
      names_from = c(mth),
      values_from = c("t"),
      names_prefix = "t",
      names_sep = "" # no underscore in new col names
    )

  # add stats

  tcols <- paste0("t", 1:12)
  t_stats_dt <- t_wide_dt |>
    mutate(
      tAvg = (t1 + t2 + t3 + t4 + t5 + t6 + t7 + t8 + t9 + t10 + t11 + t12) /
        12,
      tMax = pmax(t1, t2, t3, t4, t5, t6, t7, t8, t9, t10, t11, t12),
      tMin = pmin(t1, t2, t3, t4, t5, t6, t7, t8, t9, t10, t11, t12),
      tOctMar = (t1 + t2 + t3 + t10 + t11 + t12) / 6,
      tAprSep = (t4 + t5 + t6 + t7 + t8 + t9) / 6,
      tG10 = rowSums(across(all_of(tcols), ~ .x >= 10))
    )
  t_stats_df <- t_stats_dt |>
    select(-t1, -t2, -t3, -t4, -t5, -t6, -t7, -t8, -t9, -t10, -t11, -t12)

  return(t_stats_df)
}

```

```{r transform-p-data}

fn_transform_p_data <- function(p_dt, typ) {
  yrs <- ifelse(typ == "crp", 30, 10)
  # sort, convert to period avg, convert units

  setorder(p_dt, lon, lat, mth)
  setDT(p_dt)
  # convert from sum to avg
  p_dt[, p := p / yrs]
  # add number of days using lookup table
  p_dt[ndy, ndy := i.ndays, on = "mth"]
  p_dt[, p := p * ndy * 1000]
  p_dt[, ndy := NULL]

  # pivot

  p_wide_dt <- p_dt |>
    pivot_wider(
      names_from = c(mth),
      values_from = c("p"),
      names_prefix = "p",
      names_sep = "" # no underscore in new col names
    )

  # add stats
      
  pcols <- paste0("p", 1:12)
  p_stats_dt <- p_wide_dt |>
    mutate(
      pTot = (p1 + p2 + p3 + p4 + p5 + p6 + p7 + p8 + p9 + p10 + p11 + p12),
      pMax = pmax(p1, p2, p3, p4, p5, p6, p7, p8, p9, p10, p11, p12),
      pMin = pmin(p1, p2, p3, p4, p5, p6, p7, p8, p9, p10, p11, p12),
      pAvg = pTot / 12,
      pOctMar = (p1 + p2 + p3 + p10 + p11 + p12),
      pAprSep = (p4 + p5 + p6 + p7 + p8 + p9),
      pMinOctMar = pmin(p10, p11, p12, p1, p2, p3),
      pMaxOctMar = pmax(p10, p11, p12, p1, p2, p3),
      pMinAprSep = pmin(p4, p5, p6, p7, p8, p9),
      pMaxAprSep = pmax(p4, p5, p6, p7, p8, p9),
      pL60 = rowSums(across(all_of(pcols), ~ .x < 60))
    )
  p_stats_df <- p_stats_dt |>
    select(-p1, -p2, -p3, -p4, -p5, -p6, -p7, -p8, -p9, -p10, -p11, -p12)

  return(p_stats_df)

}

```

### Calculate grid cell area

```{r calc-grid-area}

fn_calc_area <- function(df) {
  earth_radius_km <- 6371
  lat_grid_degrees <- 0.25
  lon_grid_degrees <- 0.25
  area_df <- df |>
    mutate(
      lat_radians = lat * pi / 180,
      lat_grid_radians = lat_grid_degrees * pi / 180,
      lon_grid_radians = lon_grid_degrees * pi / 180,
      # cell area in km^2
      area = (earth_radius_km^2 *
        lon_grid_radians *
        (sin(lat_radians + lat_grid_radians / 2) -
          sin(lat_radians - lat_grid_radians / 2))) /
        1e6
    ) |>
    select(-lat_radians, -lat_grid_radians, -lon_grid_radians) |>
    select(lat, lon, area, lsm, everything())
  return(area_df)

}
```

### Save climate period data

```{r transform-crp-data}

typ <- "crp"
dir <- dir_transform
pth <- paste(dir_transform, "climate-stats-",typ,".parquet", sep = "")

# only need to transform if stats file doesn't exist
if (!file.exists(pth)) {
  t_dt <- fn_get_t_data(dir_download, typ, crp_beg, crp_end)
  p_dt <- fn_get_p_data(dir_download, typ, crp_beg, crp_end, ndy)
  t_stats_df <- fn_transform_t_data(t_dt, typ)
  p_stats_df <- fn_transform_p_data(p_dt, typ)

  j1_df <- inner_join(t_stats_df, p_stats_df, by = c("lon", "lat"))
  j2_df <- inner_join(j1_df, land_sea_mask_df, by = c("lon", "lat"))
  area_df <- fn_calc_area(j2_df)

  write_parquet(area_df, pth)
  }
```

```{r transform-dcn-data}

typ <- "dcn"
dir <- dir_transform
pth <- paste(dir_transform, "climate-stats-",typ,".parquet", sep = "")

# only need to transform if stats file doesn't exist
if (!file.exists(pth)) {
  t_dt <- fn_get_t_data(dir_download, typ, dcn_beg, dcn_end)
  p_dt <- fn_get_p_data(dir_download, typ, dcn_beg, dcn_end, ndy)
  t_stats_df <- fn_transform_t_data(t_dt, typ)
  p_stats_df <- fn_transform_p_data(p_dt, typ)

  j1_df <- inner_join(t_stats_df, p_stats_df, by = c("lon", "lat"))
  j2_df <- inner_join(j1_df, land_sea_mask_df, by = c("lon", "lat"))
  area_df <- fn_calc_area(j2_df)

  write_parquet(area_df, pth)
  }
```

### Read climate period data

```{r read-climate-period-data}
typ <- "crp"
dir <- dir_transform
pth <- paste(dir_transform, "climate-stats-",typ,".parquet", sep = "")
c_df <- read_parquet(pth)

typ <- "dcn"
dir <- dir_transform
pth <- paste(dir_transform, "climate-stats-",typ,".parquet", sep = "")
d_df <- read_parquet(pth)
```

## Exploratory Plots

### Temperature

```{r exploratory-temperature-crp}

xt1 <- ggplot(c_df) +
  geom_raster(aes(x = lon, y = lat, fill = tAvg)) +
  geom_sf(data = coastline_sf, color = "black", linewidth = 0.1) +
  scale_fill_viridis_c(option = "plasma") +
  coord_sf() +
  labs(
    title = "Climate Reference Avg Temperature",
    fill = "°C",
    x = "Longitude",
    y = "Latitude"
  ) +
  theme_minimal()
xt1
```

```{r exploratory-temperature-dcn}

xt2 <- ggplot(d_df) +
  geom_raster(aes(x = lon, y = lat, fill = tAvg)) +
  geom_sf(data = coastline_sf, color = "black", linewidth = 0.1) +
  scale_fill_viridis_c(option = "plasma") +
  coord_sf() +
  labs(
    title = "Recent Climate Avg Temperature",
    fill = "°C",
    x = "Longitude",
    y = "Latitude"
  ) +
  theme_minimal()
xt2
```

### Precipitation

```{r exploratory-precipitation-crp}
xp1 <- ggplot(c_df) +
  geom_raster(aes(x = lon, y = lat, fill = pAvg)) +
  geom_sf(data = coastline_sf, color = "black", linewidth = 0.1) +
  scale_fill_gradient(
   trans = "log10",
    low = "white", high = "blue",
    labels = function(x) round(x)) +
  coord_sf() +
  labs(
    title = "Climate Reference Avg Precipitation",
    fill = "mm",
    x = "Longitude",
    y = "Latitude"
  ) +
  theme_minimal()
xp1
        
```

```{r exploratory-precipitation-dcn}
xp2 <- ggplot(d_df) +
  geom_raster(aes(x = lon, y = lat, fill = pAvg)) +
  geom_sf(data = coastline_sf, color = "black", linewidth = 0.1) +
  scale_fill_gradient(
   trans = "log10",
    low = "white", high = "blue",
    labels = function(x) round(x)) +
  coord_sf() +
  labs(
    title = "Recent Climate Avg Precipitation",
    fill = "mm",
    x = "Longitude",
    y = "Latitude"
  ) +
  theme_minimal()
xp2
        
```

## Classification

### KTC classification function

```{r classify_ktc}
fn_classify_ktc <- function(df) {

# Step 1: Add seasonal precip stats (Summer/Winter) based on hemisphere
#         Summer: high sun 6 months (SH: Oct–Mar, NH: Apr–Sep)
#         Winter: low  sun 6 months (NH: Oct–Mar, SH: Apr–Sep)
  df <- df |>
    mutate(
      pST = case_when(lat >= 0 ~ pAprSep, TRUE ~ pOctMar),
      pWT = case_when(lat >= 0 ~ pOctMar, TRUE ~ pAprSep),
      pMnST = case_when(lat >= 0 ~ pMinAprSep, TRUE ~ pMinOctMar),
      pMxST = case_when(lat >= 0 ~ pMaxAprSep, TRUE ~ pMaxOctMar),
      pMnWT = case_when(lat >= 0 ~ pMinOctMar, TRUE ~ pMinAprSep),
      pMxWT = case_when(lat >= 0 ~ pMaxOctMar, TRUE ~ pMaxAprSep)
    )

# Step 2: Calculate Aridity threshold
  df <- df |>
    mutate(
      aThrT = case_when(
        pTot != 0 ~ (10 * (tAvg - 10)) + (3 * (pST / pTot) * 100),
        TRUE ~ 10 * (tAvg - 10)
      )
    )

# Step 3: Assign Primary Group
  df <- df |>
    mutate(
      z1 = case_when(
        tMax < 10 ~ "F",
        pTot < (aThrT * 2) ~ "B",
        tMin >= 18 ~ "A",
        tG10 >= 8 ~ "C",
        tG10 >= 4 ~ "D",
        tG10 >= 1 ~ "E",
        TRUE ~ "X"
      ),
      zn = ""
    )

# Step 4: Assign secondary group
  df <- df |>
    mutate(
      z2 = case_when(
        z1 == "A" & pL60 <= 2 ~ "r",
        z1 == "A" ~ "w",
        z1 == "B" & pTot < aThrT ~ "W",
        z1 == "B" ~ "S",
        z1 == "C" & pWT >= pST * 3 & pMin < 30 & pTot < 890 ~ "s",
        z1 == "C" ~ "f",
        z1 == "D" & tMin > 0 ~ "o",
        z1 == "D" ~ "c",
        z1 == "F" & tMax > 0 ~ "t",
        z1 == "F" ~ "i",
        TRUE ~ ""
      ),
      zn = paste(z1, z2, sep = "")
    )
  return(df)
}
```

### Classify KTC groups

```{r ktc-crp-zones}

ktc_c_df <- fn_classify_ktc(c_df)
ktc_c_df <- ktc_c_df |> select(lon, lat, zn, area, lsm)
ktc_c_sf <- st_as_sf(ktc_c_df, coords = c("lon", "lat"), crs = 4326)

ktc_d_df <- fn_classify_ktc(d_df)
ktc_d_df <- ktc_d_df |> select(lon, lat, zn, area, lsm)
ktc_d_sf <- st_as_sf(ktc_d_df, coords = c("lon", "lat"), crs = 4326)

# combine areas by crp and dcn

ktc_cd_sf <- ktc_c_sf
ktc_cd_sf <- ktc_cd_sf |> rename(czn = zn)
ktc_cd_sf$dzn <- ktc_d_sf$zn

ktc_grp_geom_sf <- ktc_cd_sf |>
  filter(lsm==1) |>
  summarise(
    geometry = st_combine(geometry),
    mkm = sum(area),
    .by = c(czn, dzn)
  ) |>
  mutate(cz1 = substr(czn, 1, 1)) |>
  mutate(dz1 = substr(dzn, 1, 1)) |>
  mutate(pct = round(mkm / sum(mkm) * 100, 3)) |>
  select(czn, cz1, dzn, dz1, mkm, pct, geometry) |>
  arrange(czn, dzn)
ktc_grp_geom_sf <- st_set_crs(ktc_grp_geom_sf, 4326)

```


### KCC classification function

 ```{r kcc-seasonal-stats}

# fn_add_seasonal_stats_K <- function(dt) {
#   # Summer in Koppen is defined as warmer 6 month period of Oct-Mar or Apr-Sep
#   # Winter in Koppen is defined as cooler 6 month period of Oct-Mar or Apr-Sep
#   dt <- dt |>
#     mutate(
#       pSK = case_when(
#         tAprSep >= tOctMar ~ pAprSep,
#         TRUE ~ pOctMar
#       ),
#       pWK = case_when(
#         tAprSep < tOctMar ~ pAprSep,
#         TRUE ~ pOctMar
#       ),
#       pMnSK = case_when(
#         tAprSep >= tOctMar ~ pmin(p4, p5, p6, p7, p8, p9),
#         TRUE ~ pmin(p1, p2, p3, p10, p11, p12),
#       ),
#       pMxSK = case_when(
#         tAprSep >= tOctMar ~ pmax(p4, p5, p6, p7, p8, p9),
#         TRUE ~ pmax(p1, p2, p3, p10, p11, p12),
#       ),
#       pMnWK = case_when(
#         tAprSep >= tOctMar ~ pmin(p1, p2, p3, p10, p11, p12),
#         TRUE ~ pmin(p4, p5, p6, p7, p8, p9),
#       ),
#       pMxWK = case_when(
#         tAprSep >= tOctMar ~ pmax(p1, p2, p3, p10, p11, p12),
#         TRUE ~ pmax(p4, p5, p6, p7, p8, p9),
#       ),
#     )
#   return(dt)
# }

 ```

```{r kcc-precipitation-threshold}

# fn_add_precipitation_threshhold_K <- function(dt) {
#   # For Koppen, the precipitation threshold in millimeters is determined by:
#   #    multiplying the average annual temperature in Celsius by 20, then adding:
#   #      280 if 70% or more of the total precipitation is in the spring and summer months (April–September in the Northern Hemisphere, or October–March in the Southern), or
#   #      140 if 30%–70% of the total precipitation is received during the spring and summer, or
#   #       0 if less than 30% of the total precipitation is received during the spring and summer.
#   dt <- dt |>
#     mutate(
#       pThrK = case_when(
#         pSK / pTot >= 0.7 ~ tAvg * 20 + 280,
#         pSK / pTot >= 0.3 ~ tAvg * 20 + 140,
#         TRUE ~ tAvg * 20
#       )
#     )
#   return(dt)
# }

```

```{r}

# fn_kcc_letter1 <- function(dt) {
#   dt <- dt |>
#     mutate(
#       k1 = case_when(
#         tMax < 10 ~ "E",
#         pTot < pThrK ~ "B",
#         tMin > 18 ~ "A",
#         tMax > 10 & tMin >= -3 ~ "C",
#         tMax > 10 & tMin <= -3 ~ "D",
#         TRUE ~ "X"
#       ),
#       k2 = "",
#       k3 = "",
#     )
#   return(dt)
# }

```

```{r}
# fn_kcc_letter2 <- function(dt) {
#   dt <- dt |>
#     mutate(
#       k2 = case_when(
#         k1 == "A" & pMin >= 60 ~ "f",
#         k1 == "A" & pMin < 60 & pMin >= (100 - pTot / 25) ~ "m",
#         k1 == "A" & pMin < (100 - pTot / 25) & pWK > pSK ~ "w",
#         k1 == "A" & pMin < (100 - pTot / 25) & pSK > pWK ~ "s",
#         k1 == "B" & pTot <= (pThrK * 0.5) ~ "W",
#         k1 == "B" & pTot <= pThrK ~ "S",
#         k1 == "C" & pMnSK < 40 & pMxWK >= pMnSK * 3 ~ "s",
#         k1 == "C" & pMxSK >= pMnWK * 10 ~ "w",
#         k1 == "C" ~ "f",
#         k1 == "D" & pMnSK < 40 & pMxWK >= pMnSK * 3 ~ "s",
#         k1 == "D" & pMxSK >= pMnWK * 10 ~ "w",
#         k1 == "D" ~ "f",
#         k1 == "E" & tMax > 0 ~ "T",
#         k1 == "E" & tMax <= 0 ~ "F",
#         TRUE ~ " "
#       ),
#     )
#   return(dt)
# }

```

```{r}
# fn_kcc_letter3 <- function(dt) {
#   dt <- dt |>
#     mutate(
#       k3 = case_when(
#         k1 == "B" & tAvg >= 18 ~ "h",
#         k1 == "B" & tAvg < 18 ~ "k",
#         k1 == "C" & tMax >= 22 ~ "a",
#         k1 == "C" & tG10 >= 4 ~ "b",
#         k1 == "C" ~ "c",
#         k1 == "D" & tMax >= 22 ~ "a",
#         k1 == "D" & tG10 >= 4 ~ "b",
#         k1 == "D" & tMin < -38 ~ "d",
#         k1 == "D" ~ "c",
#         TRUE ~ ""
#       ),
#       k12 = paste(k1, k2, sep = ""),
#       k123 = paste(k1, k2, k3, sep = "")
#     )
#   return(dt)
# }

```

### Classify KCC groups

```{r kcc-crp-zones}

# fn_kcc <- function(dt) {
#   dt <- fn_add_seasonal_stats_K(dt)
#   dt <- fn_add_precipitation_threshhold_K(dt)
#   dt <- fn_kcc_letter1(dt)
#   dt <- fn_kcc_letter2(dt)
#   dt <- fn_kcc_letter3(dt)
#   return(dt)
# }

# fn_kcc <- function(dt) {
#   dt <- fn_add_seasonal_stats_K(dt)
#   dt <- fn_add_precipitation_threshhold_K(dt)
#   dt <- fn_kcc_letter1(dt)
#   dt <- fn_kcc_letter2(dt)
#   dt <- fn_kcc_letter3(dt)
#   return(dt)
# }

```


## Summary

### KTC summary

```{r build-ktc-summary}

# Build CRP and DCN aggregates from grouped land geometries
ktc_crp_df <- ktc_grp_geom_sf |>
  st_drop_geometry() |>
  select(czn, cz1, mkm, pct) |>
  summarize(mkm = sum(mkm), pct = sum(pct), .by = c(czn, cz1)) |>
  mutate(period = "crp") |>
  rename(zone = czn, z1 = cz1) |>
  arrange(zone)

ktc_dcn_df <- ktc_grp_geom_sf |>
  st_drop_geometry() |>
  select(dzn, dz1, mkm, pct) |>
  summarize(mkm = sum(mkm), pct = sum(pct), .by = c(dzn, dz1)) |>
  mutate(period = "dcn") |>
  rename(zone = dzn, z1 = dz1) |>
  arrange(zone)

# Side-by-side comparison + deltas
ktc_diff_df <- ktc_crp_df |>
  select(zone, mkm, pct) |>
  rename(cmkm = mkm, cpct = pct)

ktc_diff_df$dmkm <- ktc_dcn_df$mkm
ktc_diff_df$dpct <- ktc_dcn_df$pct

ktc_diff_df <- ktc_diff_df |>
  mutate(
    xmkm = dmkm - cmkm,
    xpct = round(dpct - cpct, 2)
  ) |>
  select(zone, cmkm, dmkm, xmkm, cpct, dpct, xpct)

ktc_diff_df <- ktc_diff_df |>
  left_join(ktc_dict, by = "zone") |>
  select(label, cmkm, dmkm, xmkm, cpct, dpct, xpct)

ktc_diff_gt <- gt(ktc_diff_df)

```

```{r display-ktc-summary}

ktc_tbl1 <- ktc_diff_gt |>
  tab_header(
    title = "Summary of Köppen-Trewartha Climate Zone Changes",
    #subtitle = paste(crp_lbl, "vs", dcn_lbl)
  ) |>
  tab_spanner(
    label = "Area (Mkm²)",
    columns = c(cmkm, dmkm, xmkm)
  ) |>
  tab_spanner(
    label = "Earth Land Area %",
    columns = c(cpct, dpct, xpct)
  ) |>
  cols_label(
    cmkm = "CRP",
    dmkm = "DCN",
    xmkm = "Chg",
    cpct = "CRP",
    dpct = "DCN",
    xpct = "Chg"
  ) |>
  fmt_number(columns = c(cmkm, dmkm, xmkm), decimals = 1) |>
  fmt_number(columns = c(cpct, dpct, xpct), decimals = 2) |>
  grand_summary_rows(
    columns = c(cmkm, dmkm, xmkm, cpct, dpct, xpct),
    fns = list("Total" = ~ sum(.)),
    fmt = list(~ fmt_number(.x, decimals = 2))
  ) |>
  tab_footnote(
    footnote = paste("CRP: Climate Reference Period ", crp_beg, "-", crp_end, sep = ""),
    locations = cells_column_labels(columns = c(cmkm, cpct))
  ) |>
  tab_footnote(
    footnote = paste("DCN: Decadal Climate Normals ", dcn_beg, "-", dcn_end, sep = ""),
    locations = cells_column_labels(columns = c(dmkm, dpct))
  )

ktc_tbl1 
```

### KCC summary

## Visualization

### Map 1

```{r map-1}

plot_data <- ktc_grp_geom_sf |>
  select(czn) |>
  rename(zone = czn)
p1 <- ggplot(plot_data) +
  geom_sf(aes(color = zone), size = 0.1) +
  scale_color_manual(
    values = setNames(ktc_palette$color, ktc_palette$zone),
    labels = setNames(ktc_dict$label, ktc_dict$zone)
  ) +
  geom_sf(
    data = coastline_sf,
    color = "darkgrey",
    linewidth = 0.1
  ) +
  coord_sf(crs = rbn_prj) +
  labs(
    title = "Köppen-Trewartha Climate Classification",
    subtitle = crp_lbl,
    color = "Zones"
  ) +
  guides(
    color = guide_legend(
      override.aes = list(
        size = 3 # legend point size only
      )
    )
  )
p1

```

### Map 2

```{r map-2}

plot_data <- ktc_grp_geom_sf |>
  select(dzn) |>
  rename(zone = dzn)
p2 <- ggplot(plot_data) +
  geom_sf(aes(color = zone), size = 0.1) +
  scale_color_manual(
    values = setNames(ktc_palette$color, ktc_palette$zone),
    labels = setNames(ktc_dict$label, ktc_dict$zone)
  ) +
  geom_sf(
    data = coastline_sf,
    color = "darkgrey",
    linewidth = 0.1
  ) +
  coord_sf(crs = rbn_prj) +
  labs(
    title = "Köppen-Trewartha Climate Classification",
    subtitle = dcn_lbl,
    color = "Zones"
  ) +
  guides(
    color = guide_legend(
      override.aes = list(
        size = 3 # legend point size only
      )
    )
  )
p2

```

### Map 3

```{r map-3}

plot_z1 <- "B"
plot_data <- ktc_grp_geom_sf |>
  filter(cz1 != dz1) |>
  filter(dz1 == plot_z1) |>
  select(czn) |>
  rename(zone = czn)
p3 <- ggplot(plot_data) +
  geom_sf(aes(color = zone), size = 0.1) +
  scale_color_manual(
    values = setNames(ktc_palette$color, ktc_palette$zone),
    labels = setNames(ktc_dict$label, ktc_dict$zone)
  ) +
  geom_sf(
    data = coastline_sf,
    color = "darkgrey",
    linewidth = 0.1
  ) +
  coord_sf(crs = rbn_prj) +
  labs(
    title = "Köppen-Trewartha Climate Classification",
    subtitle = paste0(
      "New Type ",
      plot_z1,
      " Zones in ",
      dcn_lbl
    ),
    color = "Source Zones"
  ) +
  guides(
    color = guide_legend(
      override.aes = list(
        size = 4 # legend point size only
      )
    )
  )
p3

```

### Map 4

```{r map-4}

plot_zn <- "Cs"
plot_data <- ktc_grp_geom_sf |>
  filter(dzn == plot_zn) |>
  rename(zone = dzn)

p4 <- ggplot(plot_data) +
  geom_sf(aes(color = zone), size = 0.1) +
  scale_color_manual(
    values = setNames(ktc_palette$color, ktc_palette$zone),
    labels = setNames(ktc_dict$label, ktc_dict$zone)
  ) +
  geom_sf(
    data = coastline_sf,
    color = "darkgrey",
    linewidth = 0.1
  ) +
  coord_sf(crs = rbn_prj) +
  labs(
    title = "Köppen-Trewartha Climate Classification",
    subtitle = dcn_lbl,
    color = "Zones"
  ) +
  guides(
    color = guide_legend(
      override.aes = list(
        size = 4 # legend point size only
      )
    )
  )
p4

```
